---
title: "Audio Workflows with Microsoft Foundry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Audio Workflows with Microsoft Foundry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This vignette runs against recorded, credential-free API fixtures. The
# fixtures are recorded once by a maintainer with real Azure Speech credentials
# (data-raw/record-doc-outputs.R) and committed under vignettes/audio/. When the
# fixtures are present every foundry_*() call below executes and shows its real
# output; when they are absent the API chunks are not evaluated so the vignette
# still builds anywhere without credentials. No output on this page is fabricated.
fixture_dir <- "audio"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
run_api <- recording || have_fixtures

# Attach foundryR before start_vignette(): httptest2 only sources the package's
# inst/httptest2/start-vignette.R (which sets replay placeholders) from attached
# packages.
library(foundryR)

if (run_api) {
  httptest2::start_vignette(fixture_dir)
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_api
)
```

```{r library, eval = TRUE}
library(foundryR)
```

Audio is one of the most useful Foundry additions for researchers. You can
transcribe interviews, lectures, field recordings, meetings, and focus groups,
then keep the result in a tibble with segment-level timing for downstream coding
or analysis.

## Configure Speech in Foundry Tools

MAI-Transcribe models are exposed through the LLM Speech API. Configure the
Speech resource once per session. This chunk is illustrative and is not run when
the vignette builds:

```{r configure-speech, eval = FALSE}
foundry_set_speech_endpoint(Sys.getenv("AZURE_FOUNDRY_SPEECH_ENDPOINT"))
foundry_set_speech_key("your-speech-key")

# For keyless authentication, set a bearer token instead of a key:
foundry_set_token("your-entra-id-access-token")
```

## A real, public-domain sample

The examples below use a short excerpt from John F. Kennedy's 1961 inaugural
address ("And so, my fellow Americans..."). This clip ships with the package and
is the de facto "hello, world" of open-source speech recognition, so the
transcript is easy to check against a recording everyone knows.

```{r sample}
sample_audio <- system.file("extdata/samples/jfk.wav", package = "foundryR")
basename(sample_audio)
```

## Transcribe an audio file

`foundry_transcribe()` returns one row per file. The `text` column holds the
transcript and the `phrases` list-column holds segment-level timing.

```{r transcribe}
transcript <- foundry_transcribe(
  sample_audio,
  model = "mai-transcribe-1.5",
  locales = "en-US"
)

transcript$text
```

The segment timing lives in the `phrases` list-column, one row per recognized
phrase:

```{r transcribe-phrases}
head(transcript$phrases[[1]])
```

## Synthesize speech

`foundry_speak()` writes binary audio to disk and returns the file path and byte
count. This is useful for experiment stimuli, accessibility assets, and demos.
Use your text-to-speech deployment name for `model`.

```{r speak}
speech <- foundry_speak(
  "Welcome to the foundryR audio vignette.",
  model = "tts-1",
  voice = "alloy",
  path = tempfile(fileext = ".mp3")
)

speech[, c("bytes", "model", "voice", "format")]
```

## Translate multilingual recordings

Use `foundry_translate_audio()` when you want an analysis corpus in a common
language. To keep this example fully reproducible we first synthesize a short
Spanish clip, then translate it back to English -- both are real API calls.

```{r translate}
spanish_clip <- foundry_speak(
  "La reunion de investigacion fue clara y muy util.",
  model = "tts-1",
  voice = "alloy",
  path = tempfile(fileext = ".mp3")
)

translation <- foundry_translate_audio(
  spanish_clip$path,
  target_language = "en"
)

translation$text
```

## Notes for researchers

- Inspect `head(transcript$phrases[[1]])` before processing long recordings so
  you know the segment structure your coding scheme has to handle.
- Keep raw audio out of your project repository; store transcripts and IDs.
- MAI-Transcribe preview limitations include no diarization for MAI-Transcribe
  models and no prompt tuning for MAI-Transcribe 1.0.

```{r cleanup, include = FALSE}
if (run_api) {
  httptest2::end_vignette()
}
```
