---
title: "Audio Workflows with Microsoft Foundry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Audio Workflows with Microsoft Foundry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(foundryR)
```

Audio is one of the most useful Foundry additions for researchers. You can
transcribe interviews, lectures, field recordings, meetings, and focus groups,
then keep the result in a tibble with segment-level timing for downstream
coding or analysis.

## Configure Speech in Foundry Tools

MAI-Transcribe models are exposed through the LLM Speech API. Configure the
Speech resource once per session:

```{r configure-speech, eval = FALSE}
foundry_set_speech_endpoint("AZURE_FOUNDRY_SPEECH_ENDPOINT")
foundry_set_speech_key("your-speech-key")
```

For keyless authentication, set a bearer token instead of a key:

```{r configure-token, eval = FALSE}
foundry_set_token("your-entra-id-access-token")
```

## Transcribe an audio file

```{r transcribe-call, eval = FALSE}
transcript <- foundry_transcribe(
  "interview.mp3",
  model = "mai-transcribe-1.5",
  locales = "en-US",
  phrase_list = c("foundryR", "tidymodels")
)
```

A successful transcription returns one row per file. The `phrases` column is a
list-column with segment timing.

```{r transcribe-output}
phrases <- tibble::tibble(
  text = c(
    "We used foundryR to transcribe the interview.",
    "Then we coded the responses in R."
  ),
  locale = c("en-us", "en-us"),
  offset_ms = c(0L, 3280L),
  duration_ms = c(3120L, 2600L),
  confidence = c(0, 0),
  speaker = c(NA_character_, NA_character_),
  words = list(list(), list())
)

transcript <- tibble::tibble(
  file = "interview.mp3",
  task = "transcribe",
  model = "mai-transcribe-1.5",
  text = paste(phrases$text, collapse = " "),
  duration_ms = 5880L,
  language = "en-us",
  phrases = list(phrases)
)

transcript[, c("file", "task", "model", "text", "duration_ms")]
head(transcript$phrases[[1]])
```

## Translate multilingual recordings

Use `foundry_translate_audio()` when you want an analysis corpus in a common
language.

```{r translate-call, eval = FALSE}
translation <- foundry_translate_audio(
  "interview-es.mp3",
  target_language = "en",
  prompt = "Use concise research-note style."
)
```

```{r translate-output}
translation <- tibble::tibble(
  file = "interview-es.mp3",
  task = "translate",
  model = NA_character_,
  text = "The participant described the workflow as fast and easy to learn.",
  duration_ms = 4210L,
  language = "en",
  phrases = list(tibble::tibble(
    text = "The participant described the workflow as fast and easy to learn.",
    locale = "en",
    offset_ms = 0L,
    duration_ms = 4210L,
    confidence = 0,
    speaker = NA_character_,
    words = list(list())
  ))
)

translation[, c("file", "task", "text", "duration_ms")]
```

## Synthesize speech

`foundry_speak()` writes binary audio to disk and returns the file path and byte
count. This is useful for experiment stimuli, accessibility assets, and demos.

```{r speak-call, eval = FALSE}
speech <- foundry_speak(
  "Welcome to the study.",
  model = "tts-1",
  voice = "alloy",
  path = "welcome.mp3"
)
```

```{r speak-output}
speech <- tibble::tibble(
  path = "welcome.mp3",
  bytes = 18432L,
  model = "tts-1",
  voice = "alloy",
  format = "mp3"
)

speech
```

## Notes for researchers

- Use `head(transcript$phrases[[1]])` before processing long recordings.
- Keep raw audio out of your project repository; store transcripts and IDs.
- MAI-Transcribe preview limitations include no diarization for MAI-Transcribe
  models and no prompt tuning for MAI-Transcribe 1.0.
