---
title: "Working with Embeddings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This vignette runs against recorded, credential-free API fixtures. When the
# fixtures are present (they are recorded once with data-raw/record-doc-outputs.R
# and committed under vignettes/embeddings/), every foundry_*() call below is
# executed and its real output is shown. When they are absent, the API chunks are
# not evaluated so the vignette still builds anywhere without Azure credentials.
fixture_dir <- "embeddings"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
run_api <- requireNamespace("httptest2", quietly = TRUE) &&
  (recording || have_fixtures)

# Attach foundryR before start_vignette(): httptest2 only sources the package's
# inst/httptest2/start-vignette.R (which sets replay placeholders) from attached
# packages.
library(foundryR)

if (run_api) {
  httptest2::start_vignette(fixture_dir)
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_api
)
```

## What embeddings are for

Embeddings are numerical representations of text that capture semantic meaning.
When you convert text to an embedding, you get a vector of numbers (often 1,536
or 3,072 dimensions depending on the model). Texts with similar meanings tend to
have similar vectors.

Use embeddings when you need to:

- Find documents related to a query by meaning, not only keywords.
- Measure how similar two pieces of text are.
- Cluster open-ended responses into themes.
- Find near-duplicate responses or records.
- Feed text-derived numeric predictors into downstream models.

Unlike keyword matching, embeddings understand that "automobile" and "car" are
semantically similar, even though they share no letters.

## Generating embeddings with foundry_embed()

The examples below embed real sentences from Jane Austen's *Pride and Prejudice*,
available in the `janeaustenr` package. Using a well-known public-domain text
makes the output easy to reason about: the opening lines share vocabulary and
sentiment, so their embeddings should sit close together.

```{r austen-lines}
library(foundryR)

austen_lines <- c(
  "It is a truth universally acknowledged, that a single man in possession of a good fortune, must be in want of a wife.",
  "However little known the feelings or views of such a man may be on his first entering a neighbourhood.",
  "Mr. Bennet was so odd a mixture of quick parts, sarcastic humour, reserve, and caprice."
)

embedding <- foundry_embed(austen_lines[1], model = "text-embedding-3-small")
embedding
```

The result is a tibble with:

- `text`: the original input text.
- `embedding`: a list-column holding the numeric vector.
- `n_dims`: the dimensionality of the embedding.

### Embedding multiple texts

Pass a character vector to embed several texts in one call:

```{r multiple-embed}
doc_embeddings <- foundry_embed(austen_lines, model = "text-embedding-3-small")
doc_embeddings
```

### Controlling dimensions

The `text-embedding-3-small` and `text-embedding-3-large` models can return
shorter vectors. Smaller dimensions mean faster similarity computations and less
storage, with some trade-off in precision:

```{r reduced-dims}
compact <- foundry_embed(
  austen_lines[1],
  model = "text-embedding-3-small",
  dimensions = 256
)
compact$n_dims
```

## Computing similarity with foundry_similarity()

Cosine similarity measures how close two embeddings are, from -1 (opposite) to 1
(identical). `foundry_similarity()` computes every pairwise similarity in a tibble
of embeddings. Here we contrast the Austen lines with two sentences from a
different domain so the split is visible.

```{r similarity}
mixed <- c(
  "It is a truth universally acknowledged, that a single man in possession of a good fortune, must be in want of a wife.",
  "Mr. Bennet was so odd a mixture of quick parts, sarcastic humour, reserve, and caprice.",
  "The quarterly revenue report showed a sharp rise in cloud subscriptions.",
  "Analysts raised their earnings forecast after the strong cloud numbers."
)

similarities <- foundry_embed(mixed, model = "text-embedding-3-small") |>
  foundry_similarity()
similarities
```

Results are sorted by similarity. The two Austen lines pair together and the two
finance lines pair together, while cross-domain pairs score lower.
The plots in this vignette are shown when the suggested `ggplot2` package is
installed.

```{r similarity-pairs-data, include = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE)}
label_for <- function(x) {
  dplyr::case_when(
    startsWith(x, "It is a truth") ~ "Austen 1",
    startsWith(x, "Mr. Bennet") ~ "Austen 2",
    startsWith(x, "The quarterly") ~ "Finance 1",
    TRUE ~ "Finance 2"
  )
}

pair_data <- similarities |>
  dplyr::mutate(
    first = pmin(label_for(text_1), label_for(text_2)),
    second = pmax(label_for(text_1), label_for(text_2)),
    pair = paste(first, "and", second),
    source_match = ifelse(
      sub(" .*$", "", first) == sub(" .*$", "", second),
      "same",
      "different"
    )
  ) |>
  dplyr::arrange(similarity)
stopifnot(!anyDuplicated(pair_data$pair))
pair_data$pair <- factor(pair_data$pair, levels = pair_data$pair)

same_scores <- pair_data$similarity[pair_data$source_match == "same"]
cross_scores <- pair_data$similarity[pair_data$source_match == "different"]
similarity_chart_title <- if (min(same_scores) > max(cross_scores)) {
  "Every same-source pair scores higher than every cross-source pair"
} else {
  "Same-source pairs tend to score higher than cross-source pairs"
}
similarity_chart_alt <- paste0(
  similarity_chart_title,
  ". Horizontal bar chart of cosine similarity for six sentence pairs: ",
  paste(
    sprintf("%s %.2f", rev(as.character(pair_data$pair)), rev(pair_data$similarity)),
    collapse = "; "
  ),
  "."
)
```

```{r similarity-pairs, echo = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE), fig.alt = get0("similarity_chart_alt", ifnotfound = "Bar chart of cosine similarity for sentence pairs from two sources.")}
ggplot2::ggplot(
  pair_data,
  ggplot2::aes(x = similarity, y = pair, fill = source_match)
) +
  ggplot2::geom_col(width = 0.66) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.2f", similarity), color = source_match),
    hjust = -0.25,
    size = 3.6
  ) +
  ggplot2::scale_fill_manual(
    values = c(same = "#0072B2", different = "#A6A6A6"),
    guide = "none"
  ) +
  ggplot2::scale_color_manual(
    values = c(same = "#005A8C", different = "#595959"),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    expand = ggplot2::expansion(mult = c(0, 0.02))
  ) +
  ggplot2::coord_cartesian(xlim = c(min(0, pair_data$similarity), 1)) +
  ggplot2::labs(
    title = similarity_chart_title,
    subtitle = "Cosine similarity for all six pairs of two Austen lines and two finance sentences",
    x = "Cosine similarity",
    y = NULL,
    caption = "Blue bars pair sentences from the same source. Scores from foundry_similarity() on text-embedding-3-small embeddings."
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle = ggplot2::element_text(color = "grey30"),
    plot.caption = ggplot2::element_text(color = "grey40", hjust = 0),
    plot.caption.position = "plot",
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(color = "grey15")
  )
```

## Use case: finding similar documents

A common application is ranking documents by relevance to a query. Embed the
documents and the query, then sort by cosine similarity:

```{r semantic-search}
library(dplyr)

documents <- c(
  "How to install R packages using install.packages()",
  "Data visualization with ggplot2 in R",
  "Introduction to machine learning with Python",
  "Statistical hypothesis testing explained",
  "Building web applications with Shiny",
  "Deep learning with TensorFlow and Keras"
)

doc_embeddings <- foundry_embed(documents, model = "text-embedding-3-small")
query_embedding <- foundry_embed(
  "How do I create charts and graphs in R?",
  model = "text-embedding-3-small"
)

cosine <- function(a, b) sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))
query_vec <- query_embedding$embedding[[1]]

doc_embeddings |>
  mutate(similarity = vapply(embedding, cosine, numeric(1), b = query_vec)) |>
  arrange(desc(similarity)) |>
  select(text, similarity) |>
  head(3)
```

## Use case: clustering text

Embeddings work well as features for clustering. Here `stats::kmeans()` groups a
mix of programming, food, and sports sentences without any labels:

```{r clustering}
texts <- c(
  "Python is great for machine learning",
  "R excels at statistical analysis",
  "JavaScript powers modern web applications",
  "Italian pasta with tomato sauce",
  "Sushi is a popular Japanese dish",
  "French croissants are flaky and buttery",
  "Soccer is the world's most popular sport",
  "Basketball requires speed and agility",
  "Tennis matches can last for hours"
)

cluster_embeddings <- foundry_embed(texts, model = "text-embedding-3-small")
embedding_matrix <- do.call(rbind, cluster_embeddings$embedding)

set.seed(42)
clusters <- kmeans(embedding_matrix, centers = 3, nstart = 10)

cluster_embeddings |>
  mutate(cluster = clusters$cluster) |>
  arrange(cluster) |>
  select(text, cluster)
```

```{r projection-data, include = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE)}
topics <- rep(c("Programming", "Food", "Sports"), each = 3)
stopifnot(length(topics) == length(texts))

pca <- stats::prcomp(embedding_matrix, rank. = 2)
projection <- tibble::tibble(
  pc1 = pca$x[, 1],
  pc2 = pca$x[, 2],
  cluster = factor(clusters$cluster),
  topic = topics
)

cluster_labels <- projection |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(
    n_points = dplyr::n(),
    top_count = max(table(topic)),
    label_text = names(which.max(table(topic))),
    label_x = mean(pc1),
    label_y = max(pc2),
    .groups = "drop"
  )
label_offset <- 0.08 * diff(range(projection$pc2))

cluster_colors <- c("#0072B2", "#D55E00", "#009E73")
cluster_text <- c("#005A8C", "#A84A0A", "#00735A")
cluster_shapes <- c(16, 17, 15)

clusters_are_pure <- all(cluster_labels$top_count == cluster_labels$n_points) &&
  dplyr::n_distinct(cluster_labels$label_text) == nrow(cluster_labels)
projection_title <- if (clusters_are_pure) {
  "k-means recovers the three topics from the embeddings alone"
} else {
  "k-means clusters mostly follow the three topics"
}
projection_alt <- paste0(
  projection_title,
  ". Scatter plot of nine sentence embeddings on their first two principal components; ",
  paste(
    sprintf(
      "cluster %s holds %d of %d %s sentences",
      cluster_labels$cluster,
      cluster_labels$top_count,
      cluster_labels$n_points,
      tolower(cluster_labels$label_text)
    ),
    collapse = "; "
  ),
  "."
)
```

```{r projection, echo = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE), fig.alt = get0("projection_alt", ifnotfound = "Scatter plot of sentence embeddings projected onto two principal components.")}
ggplot2::ggplot(
  projection,
  ggplot2::aes(pc1, pc2, color = cluster, shape = cluster)
) +
  ggplot2::geom_point(size = 3.4) +
  ggplot2::geom_text(
    data = cluster_labels,
    ggplot2::aes(x = label_x, y = label_y + label_offset, label = label_text),
    inherit.aes = FALSE,
    color = cluster_text[as.integer(cluster_labels$cluster)],
    fontface = "bold",
    size = 4.2,
    vjust = 0
  ) +
  ggplot2::scale_color_manual(values = cluster_colors, guide = "none") +
  ggplot2::scale_shape_manual(values = cluster_shapes, guide = "none") +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = 0.12)) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.08, 0.22))) +
  ggplot2::coord_cartesian(clip = "off") +
  ggplot2::labs(
    title = projection_title,
    subtitle = "Nine sentences on the first two principal components of their embeddings;\ncolor and shape show the k-means cluster (k = 3)",
    x = "Principal component 1",
    y = "Principal component 2",
    caption = "Embeddings from text-embedding-3-small via foundry_embed(). Axis positions are relative, so tick values are omitted."
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle = ggplot2::element_text(color = "grey30"),
    plot.caption = ggplot2::element_text(color = "grey40", hjust = 0),
    plot.caption.position = "plot",
    axis.text = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank()
  )
```

The clusters recover the three topics from the raw text alone.

## Tips for working with embeddings

### Choosing a model

| Model | Dimensions | Notes |
|-------|-----------|-------|
| text-embedding-ada-002 | 1,536 | Previous generation, widely used |
| text-embedding-3-small | 1,536 (configurable) | Newer, supports dimension reduction |
| text-embedding-3-large | 3,072 (configurable) | Highest quality, more expensive |

For most use cases, `text-embedding-3-small` balances quality and cost.

### Dimension trade-offs

Higher dimensions capture more nuance but need more storage, take longer to
compare, and may not improve simple tasks. Consider reduced dimensions (256-512)
for large-scale applications where speed matters more than precision.

### Handling large collections

1. **Batch processing**: embed documents in batches to respect rate limits.
2. **Caching**: store embeddings in a database rather than regenerating them.
3. **Approximate nearest neighbors**: use libraries like `RcppAnnoy` for fast
   similarity search on large datasets.

```{r batch-example, eval = TRUE}
# Defining this helper is local; calling it requires Azure credentials.
batch_embed <- function(texts, model, batch_size = 100) {
  n_batches <- ceiling(length(texts) / batch_size)
  results <- vector("list", n_batches)
  for (i in seq_len(n_batches)) {
    start_idx <- (i - 1) * batch_size + 1
    end_idx <- min(i * batch_size, length(texts))
    results[[i]] <- foundry_embed(texts[start_idx:end_idx], model = model)
    Sys.sleep(0.5)
  }
  dplyr::bind_rows(results)
}
```

```{r cleanup, include = FALSE}
if (run_api) {
  httptest2::end_vignette()
}
```
