---
title: "Embeddings for research"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Embeddings for research}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
fixture_dir <- "embeddings"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
run_api <- requireNamespace("httptest2", quietly = TRUE) &&
  (recording || have_fixtures)
library(foundryR)
if (run_api) {
  httptest2::start_vignette(fixture_dir)
}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = run_api,
  fig.width = 7, fig.height = 4.5, out.width = "100%")
```

Calls to Azure show output recorded from a live run, and setup code is shown but not run.

## What embeddings are for

Embeddings are numeric vectors that represent text. Texts with similar meanings tend to have vectors that point in similar directions, so cosine similarity becomes a useful way to compare responses, documents, or queries.

Use embeddings when you need to find related documents by meaning, measure similarity between texts, cluster open-ended responses, detect near duplicates, or turn text into numeric predictors for a model. Keyword matching still has value when exact terms matter. Embeddings help when two texts use different wording for a similar idea, such as "automobile" and "car".

## Generate embeddings

`foundry_embed()` returns one row per input text. The `embedding` column is a list-column of numeric vectors, and `n_dims` reports the length of each vector. Metadata columns keep the original input index, row-level errors, error messages, and per-row response metadata.

```{r libraries, message = FALSE}
library(foundryR)
library(dplyr)
```

```{r austen-lines}
austen_lines <- c(
  "It is a truth universally acknowledged, that a single man in possession of a good fortune, must be in want of a wife.",
  "However little known the feelings or views of such a man may be on his first entering a neighbourhood.",
  "Mr. Bennet was so odd a mixture of quick parts, sarcastic humour, reserve, and caprice."
)

embedding <- foundry_embed(austen_lines[1], model = "text-embedding-3-small")
embedding[, c("text", "n_dims", ".input_idx", ".error", ".error_msg")]
```

Pass a character vector to embed several texts in one request.

```{r multiple-embed}
doc_embeddings <- foundry_embed(austen_lines, model = "text-embedding-3-small")
doc_embeddings[, c("text", "n_dims", ".input_idx", ".error")]
```

Some embedding deployments can return shorter vectors. That can reduce storage and speed up comparisons, with some loss of detail.

```{r reduced-dims}
compact <- foundry_embed(
  austen_lines[1],
  model = "text-embedding-3-small",
  dimensions = 256
)
compact[, c("text", "n_dims")]
```

## Compare same-source and cross-source pairs

`foundry_similarity()` computes pairwise cosine similarity for a tibble with an `embedding` list-column. The next example mixes two Austen-style lines with two finance sentences, so the useful check is whether same-source pairs score above cross-source pairs.

```{r similarity}
mixed <- c(
  "It is a truth universally acknowledged, that a single man in possession of a good fortune, must be in want of a wife.",
  "Mr. Bennet was so odd a mixture of quick parts, sarcastic humour, reserve, and caprice.",
  "The quarterly revenue report showed a sharp rise in cloud subscriptions.",
  "Analysts raised their earnings forecast after the strong cloud numbers."
)

similarities <- foundry_embed(mixed, model = "text-embedding-3-small") |>
  foundry_similarity()
similarities
```

```{r similarity-summary, echo = FALSE, results = "asis"}
pair_score <- function(a, b) {
  hit <- (similarities$text_1 == a & similarities$text_2 == b) |
    (similarities$text_1 == b & similarities$text_2 == a)
  similarities$similarity[hit][[1]]
}
finance_pair <- pair_score(mixed[3], mixed[4])
austen_pair <- pair_score(mixed[1], mixed[2])
cross_max <- max(setdiff(similarities$similarity, c(finance_pair, austen_pair)))
cat(sprintf(
  "In this recording the two finance sentences score %.2f and the two Austen lines %.2f, while no cross-source pair scores above %.2f. The Austen lines have no content words in common, so their score comes from meaning and style rather than shared vocabulary.\n",
  finance_pair, austen_pair, cross_max
))
```

```{r similarity-pairs-data, include = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE)}
label_for <- function(x) {
  dplyr::case_when(
    startsWith(x, "It is a truth") ~ "Austen 1",
    startsWith(x, "Mr. Bennet") ~ "Austen 2",
    startsWith(x, "The quarterly") ~ "Finance 1",
    TRUE ~ "Finance 2"
  )
}

pair_data <- similarities |>
  dplyr::mutate(
    first = pmin(label_for(text_1), label_for(text_2)),
    second = pmax(label_for(text_1), label_for(text_2)),
    pair = paste(first, "and", second),
    source_match = ifelse(
      sub(" .*$", "", first) == sub(" .*$", "", second),
      "same",
      "different"
    )
  ) |>
  dplyr::arrange(similarity)
stopifnot(!anyDuplicated(pair_data$pair))
pair_data$pair <- factor(pair_data$pair, levels = pair_data$pair)

same_scores <- pair_data$similarity[pair_data$source_match == "same"]
cross_scores <- pair_data$similarity[pair_data$source_match == "different"]
similarity_chart_title <- if (min(same_scores) > max(cross_scores)) {
  "Same-source pairs score above all cross-source pairs"
} else {
  "Same-source pairs tend to score above cross-source pairs"
}
wrap_text <- function(lines, width) {
  wrapped <- vapply(
    lines,
    function(line) paste(strwrap(line, width), collapse = "\n"),
    character(1)
  )
  paste(wrapped, collapse = "\n")
}
similarity_chart_alt <- paste0(
  similarity_chart_title,
  ". Horizontal bar chart of cosine similarity for six sentence pairs: ",
  paste(
    sprintf("%s %.2f", rev(as.character(pair_data$pair)), rev(pair_data$similarity)),
    collapse = "; "
  ),
  "."
)
```

```{r similarity-pairs, echo = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE), fig.alt = get0("similarity_chart_alt", ifnotfound = "Bar chart of cosine similarity for sentence pairs from two sources.")}
ggplot2::ggplot(
  pair_data,
  ggplot2::aes(x = similarity, y = pair, fill = source_match)
) +
  ggplot2::geom_col(width = 0.66) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.2f", similarity), color = source_match),
    hjust = -0.25,
    size = 3.6
  ) +
  ggplot2::scale_fill_manual(
    values = c(same = "#0072B2", different = "#A6A6A6"),
    guide = "none"
  ) +
  ggplot2::scale_color_manual(
    values = c(same = "#005A8C", different = "#595959"),
    guide = "none"
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(0, 1, by = 0.25),
    expand = ggplot2::expansion(mult = c(0, 0.02))
  ) +
  ggplot2::coord_cartesian(xlim = c(min(0, pair_data$similarity), 1)) +
  ggplot2::labs(
    title = wrap_text(similarity_chart_title, 62),
    subtitle = wrap_text(
      "Cosine similarity for all six pairs of two Austen lines and two finance sentences",
      84
    ),
    x = "Cosine similarity",
    y = NULL,
    caption = wrap_text(
      c(
        "Blue bars pair sentences from the same source.",
        "Scores come from foundry_similarity() on text-embedding-3-small embeddings."
      ),
      101
    )
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle = ggplot2::element_text(color = "grey30"),
    plot.caption = ggplot2::element_text(color = "grey40", hjust = 0),
    plot.caption.position = "plot",
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(color = "grey15")
  )
```

## Rank documents for a query

For a query-versus-corpus comparison, combine the query and documents in one embedding tibble, compute pairwise similarities, and keep the rows that include the query label. This uses `foundry_similarity()` for the cosine calculation and only a small amount of reshaping to return document text.

```{r semantic-search}
documents <- c(
  "How to install R packages using install.packages()",
  "Data visualization with ggplot2 in R",
  "Introduction to machine learning with Python",
  "Statistical hypothesis testing explained",
  "Building web applications with Shiny",
  "Deep learning with TensorFlow and Keras"
)
query <- "How do I create charts and graphs in R?"

search_embeddings <- foundry_embed_batch(
  c(query, documents),
  model = "text-embedding-3-small",
  batch_size = 4,
  max_active = 2
) |>
  mutate(label = c("query", paste0("doc_", seq_along(documents))))

search_pairs <- foundry_similarity(search_embeddings, text_col = "label")

search_pairs |>
  filter(text_1 == "query" | text_2 == "query") |>
  mutate(
    label = if_else(text_1 == "query", text_2, text_1),
    text = documents[match(label, paste0("doc_", seq_along(documents)))]
  ) |>
  select(text, similarity) |>
  arrange(desc(similarity)) |>
  slice_head(n = 3)
```

## Cluster text responses

Embeddings also work as features for local methods such as `stats::kmeans()`. The example groups a mix of programming, food, and sports sentences without using the topic labels.

```{r clustering}
texts <- c(
  "Python is great for machine learning",
  "R excels at statistical analysis",
  "JavaScript powers modern web applications",
  "Italian pasta with tomato sauce",
  "Sushi is a popular Japanese dish",
  "French croissants are flaky and buttery",
  "Soccer is the world's most popular sport",
  "Basketball requires speed and agility",
  "Tennis matches can last for hours"
)

cluster_embeddings <- foundry_embed(texts, model = "text-embedding-3-small")
embedding_matrix <- do.call(rbind, cluster_embeddings$embedding)

set.seed(42)
clusters <- kmeans(embedding_matrix, centers = 3, nstart = 10)

cluster_embeddings |>
  mutate(cluster = clusters$cluster) |>
  select(text, cluster) |>
  arrange(cluster)
```

```{r projection-data, include = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE)}
topics <- rep(c("Programming", "Food", "Sports"), each = 3)
stopifnot(length(topics) == length(texts))

pca <- stats::prcomp(embedding_matrix, rank. = 2)
projection <- tibble::tibble(
  pc1 = pca$x[, 1],
  pc2 = pca$x[, 2],
  cluster = factor(clusters$cluster),
  topic = topics
)

cluster_labels <- projection |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(
    n_points = dplyr::n(),
    top_count = max(table(topic)),
    label_text = names(which.max(table(topic))),
    label_x = mean(pc1),
    label_y = max(pc2),
    .groups = "drop"
  )
label_offset <- 0.08 * diff(range(projection$pc2))

cluster_colors <- c("#0072B2", "#D55E00", "#009E73")
cluster_text <- c("#005A8C", "#A84A0A", "#00735A")
cluster_shapes <- c(16, 17, 15)

clusters_are_pure <- all(cluster_labels$top_count == cluster_labels$n_points) &&
  dplyr::n_distinct(cluster_labels$label_text) == nrow(cluster_labels)
projection_title <- if (clusters_are_pure) {
  "k-means separates the three topics in embedding space"
} else {
  "k-means clusters mostly follow the three topics"
}
projection_alt <- paste0(
  projection_title,
  ". Scatter plot of nine sentence embeddings on their first two principal components; ",
  paste(
    sprintf(
      "cluster %s holds %d of %d %s sentences",
      cluster_labels$cluster,
      cluster_labels$top_count,
      cluster_labels$n_points,
      tolower(cluster_labels$label_text)
    ),
    collapse = "; "
  ),
  "."
)
```

```{r projection, echo = FALSE, eval = run_api && requireNamespace("ggplot2", quietly = TRUE), fig.alt = get0("projection_alt", ifnotfound = "Scatter plot of sentence embeddings projected onto two principal components.")}
ggplot2::ggplot(
  projection,
  ggplot2::aes(pc1, pc2, color = cluster, shape = cluster)
) +
  ggplot2::geom_point(size = 3.4) +
  ggplot2::geom_text(
    data = cluster_labels,
    ggplot2::aes(x = label_x, y = label_y + label_offset, label = label_text),
    inherit.aes = FALSE,
    color = cluster_text[as.integer(cluster_labels$cluster)],
    fontface = "bold",
    size = 4.2,
    vjust = 0
  ) +
  ggplot2::scale_color_manual(values = cluster_colors, guide = "none") +
  ggplot2::scale_shape_manual(values = cluster_shapes, guide = "none") +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = 0.12)) +
  ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0.08, 0.22))) +
  ggplot2::coord_cartesian(clip = "off") +
  ggplot2::labs(
    title = wrap_text(projection_title, 62),
    subtitle = wrap_text(
      c(
        "Nine sentences on the first two principal components of their embeddings",
        "Color and shape show the k-means cluster with k = 3"
      ),
      84
    ),
    x = "Principal component 1",
    y = "Principal component 2",
    caption = wrap_text(
      c(
        "Embeddings come from text-embedding-3-small via foundry_embed().",
        "Axis positions are relative, so tick values are omitted."
      ),
      101
    )
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle = ggplot2::element_text(color = "grey30"),
    plot.caption = ggplot2::element_text(color = "grey40", hjust = 0),
    plot.caption.position = "plot",
    axis.text = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank()
  )
```

The cluster assignments are an inspection tool, not proof that the labels are correct. Review the text in each cluster before using the groups as measurements.

## Work with larger collections

Use `foundry_embed_batch()` when you have many texts. It accepts `batch_size` and `max_active`, returns the same embedding list-column shape, and records row-level errors instead of stopping the whole job for one failed text.

```{r batch-example}
many_texts <- c(
  "The login page rejects my password reset link.",
  "The invoice total does not match the purchase order.",
  "The chart export button is missing from the report.",
  "The password reset email arrived after it expired.",
  "The billing address changed but the invoice did not update.",
  "The dashboard chart uses the wrong date range."
)

many_embeddings <- foundry_embed_batch(
  many_texts,
  model = "text-embedding-3-small",
  batch_size = 3,
  max_active = 2
)

many_embeddings[, c("text", "n_dims", ".error", ".error_msg")]
```

For production use, store embeddings instead of regenerating them, keep the model and dimension settings with the stored data, and use approximate-nearest-neighbor indexes when a full pairwise comparison is too slow.

```{r cleanup, include = FALSE}
if (run_api) {
  httptest2::end_vignette()
}
```
