---
title: "Working with Embeddings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

foundryr_vignette_cache <- function(name) {
  path <- system.file("extdata", "vignette-cache", name, package = "foundryR")
  if (!nzchar(path)) {
    path <- file.path("..", "inst", "extdata", "vignette-cache", name)
  }
  path
}
```

## What embeddings are for

Embeddings are numerical representations of text that capture semantic meaning.
When you convert text to an embedding, you get a vector of numbers (often 1,536
or 3,072 dimensions depending on the model). Texts with similar meanings tend to
have similar vectors.

Use embeddings when you need to:

- Find documents related to a query by meaning, not only keywords.
- Measure how similar two pieces of text are.
- Cluster open-ended responses into themes.
- Find near-duplicate responses or records.
- Feed text-derived numeric predictors into downstream models.

Unlike keyword matching, embeddings understand that "automobile" and "car" are semantically similar, even though they share no letters.

## Generating Embeddings with foundry_embed()

The `foundry_embed()` function converts text into embedding vectors:

```{r basic-embed}
library(foundryR)

# Single text
embedding <- foundry_embed("Machine learning is transforming industries",
                           model = "my-embeddings")
embedding
#> # A tibble: 1 x 3
#>   text                                       embedding      n_dims
#>   <chr>                                      <list>          <int>
#> 1 Machine learning is transforming industries <dbl [1,536]>   1536
```

The result is a tibble with:

- `text`: The original input text
- `embedding`: A list-column containing the numeric vector
- `n_dims`: The dimensionality of the embedding

### Embedding Multiple Texts

Pass a character vector to embed multiple texts in one call:

```{r multiple-embed}
documents <- c(
  "The quick brown fox jumps over the lazy dog",
  "A fast auburn fox leaps above a sleepy canine",
  "The stock market closed higher today",
  "Financial markets saw gains in afternoon trading"
)

doc_embeddings <- foundry_embed(documents, model = "my-embeddings")
doc_embeddings
#> # A tibble: 4 x 3
#>   text                                            embedding      n_dims
#>   <chr>                                           <list>          <int>
#> 1 The quick brown fox jumps over the lazy dog     <dbl [1,536]>    1536
#> 2 A fast auburn fox leaps above a sleepy canine   <dbl [1,536]>    1536
#> 3 The stock market closed higher today            <dbl [1,536]>    1536
#> 4 Financial markets saw gains in afternoon trading <dbl [1,536]>   1536
```

### Controlling Dimensions

Some embedding models (like `text-embedding-3-small` and `text-embedding-3-large`) support reducing the output dimensions. Smaller dimensions mean faster similarity computations and less storage, with some trade-off in precision:

```{r reduced-dims}
# Reduce to 256 dimensions (model must support this)
compact_embedding <- foundry_embed(
  "Hello, world!",
  model = "my-text-embedding-3-small",
  dimensions = 256
)
compact_embedding$n_dims
#> [1] 256
```

## Computing Similarity with foundry_similarity()

Cosine similarity measures how similar two embeddings are, with values ranging from -1 (opposite) to 1 (identical). The `foundry_similarity()` function computes pairwise similarities for all embeddings in a tibble:

```{r similarity}
texts <- c(
  "I love programming in R",
  "R is my favorite language for data analysis",
  "The weather is nice today",
  "It's sunny and warm outside"
)

embeddings <- foundry_embed(texts, model = "my-embeddings")
similarities <- foundry_similarity(embeddings)
similarities
#> # A tibble: 6 x 3
#>   text_1                                   text_2                              similarity
#>   <chr>                                    <chr>                                    <dbl>
#> 1 The weather is nice today                It's sunny and warm outside              0.934
#> 2 I love programming in R                  R is my favorite language for da...      0.912
#> 3 I love programming in R                  The weather is nice today                0.723
#> 4 I love programming in R                  It's sunny and warm outside              0.698
#> 5 R is my favorite language for data an... The weather is nice today                0.687
#> 6 R is my favorite language for data an... It's sunny and warm outside              0.672
```

Results are sorted by similarity (highest first). Notice how texts about similar topics (R programming, weather) have higher similarity scores with each other.

The rendered outputs below use precomputed, illustrative data from a
representative run. They are rendered during the documentation build without
Azure credentials.

```{r similarity-heatmap-rendered, eval = TRUE, echo = FALSE, fig.alt = "Heatmap of cosine similarity across four short texts."}
embedding_cache <- readRDS(foundryr_vignette_cache("embeddings.rds"))
similarity_heatmap <- embedding_cache$similarity_heatmap |>
  dplyr::mutate(
    text_1_short = dplyr::case_when(
      text_1 == "I love programming in R" ~ "R programming",
      text_1 == "R is my favorite language for data analysis" ~ "R analysis",
      text_1 == "The weather is nice today" ~ "Weather",
      TRUE ~ "Sunny"
    ),
    text_2_short = dplyr::case_when(
      text_2 == "I love programming in R" ~ "R programming",
      text_2 == "R is my favorite language for data analysis" ~ "R analysis",
      text_2 == "The weather is nice today" ~ "Weather",
      TRUE ~ "Sunny"
    )
  )

ggplot2::ggplot(
  similarity_heatmap,
  ggplot2::aes(x = text_2_short, y = text_1_short, fill = similarity)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.6) +
  ggplot2::geom_text(
    ggplot2::aes(label = sprintf("%.2f", similarity)),
    size = 3.2
  ) +
  ggplot2::scale_fill_gradient(low = "#E6F2FB", high = "#0078D4", limits = c(0.65, 1)) +
  ggplot2::labs(
    title = "Cosine similarity separates near matches from weaker pairs",
    x = NULL,
    y = NULL,
    fill = "Similarity"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    legend.position = "bottom"
  )
```

## Use Case: Finding Similar Documents

A common application is finding documents most similar to a query. Here is how to implement a simple semantic search:

```{r semantic-search}
library(dplyr)

# Your document collection
documents <- c(
  "How to install R packages using install.packages()",
  "Data visualization with ggplot2 in R",
  "Introduction to machine learning with Python",
  "Statistical hypothesis testing explained",
  "Building web applications with Shiny",
  "Deep learning with TensorFlow and Keras"
)

# Embed all documents
doc_embeddings <- foundry_embed(documents, model = "my-embeddings")

# User query
query <- "How do I create charts and graphs in R?"
query_embedding <- foundry_embed(query, model = "my-embeddings")

# Compute similarity between query and each document
compute_similarity <- function(emb1, emb2) {
  sum(emb1 * emb2) / (sqrt(sum(emb1^2)) * sqrt(sum(emb2^2)))
}

query_vec <- query_embedding$embedding[[1]]

results <- doc_embeddings %>%
  mutate(
    similarity = sapply(embedding, function(e) compute_similarity(query_vec, e))
  ) %>%
  arrange(desc(similarity)) %>%
  select(text, similarity)

# Top 3 most relevant documents
head(results, 3)
#> # A tibble: 3 x 2
#>   text                                          similarity
#>   <chr>                                              <dbl>
#> 1 Data visualization with ggplot2 in R              0.891
#> 2 How to install R packages using install.packages() 0.812
#> 3 Building web applications with Shiny              0.756
```

```{r nearest-neighbors-rendered, eval = TRUE, echo = FALSE}
embedding_cache$nearest_neighbors |>
  dplyr::mutate(similarity = round(similarity, 3)) |>
  gt::gt() |>
  gt::cols_label(
    rank = "Rank",
    text = "Document",
    similarity = "Cosine similarity"
  ) |>
  gt::fmt_number(columns = similarity, decimals = 3) |>
  gt::tab_header(title = "Nearest documents for a charting query") |>
  gt::tab_options(table.font.names = "Inter")
```

## Use Case: Clustering Text

Embeddings work well as features for clustering algorithms. Here is an example using `stats::kmeans()` to group similar texts:

```{r clustering}
library(dplyr)

# Sample texts to cluster
texts <- c(
  # Tech/Programming cluster
  "Python is great for machine learning",
  "R excels at statistical analysis",
  "JavaScript powers modern web applications",
  "SQL is essential for database queries",
  # Food cluster
"Italian pasta with tomato sauce",
  "Sushi is a popular Japanese dish",
  "French croissants are flaky and buttery",
  "Mexican tacos with fresh salsa",
  # Sports cluster
  "Soccer is the world's most popular sport",
  "Basketball requires speed and agility",
  "Tennis matches can last for hours",
  "Swimming is excellent cardiovascular exercise"
)

# Generate embeddings
embeddings <- foundry_embed(texts, model = "my-embeddings")

# Convert list-column to matrix for kmeans
embedding_matrix <- do.call(rbind, embeddings$embedding)

# Cluster into 3 groups
set.seed(42)
clusters <- kmeans(embedding_matrix, centers = 3, nstart = 10)

# Add cluster assignments to our data
results <- embeddings %>%
  mutate(cluster = clusters$cluster) %>%
  arrange(cluster) %>%
  select(text, cluster)

results
#> # A tibble: 12 x 2
#>    text                                       cluster
#>    <chr>                                        <int>
#>  1 Python is great for machine learning             1
#>  2 R excels at statistical analysis                 1
#>  3 JavaScript powers modern web applications        1
#>  4 SQL is essential for database queries            1
#>  5 Italian pasta with tomato sauce                  2
#>  6 Sushi is a popular Japanese dish                 2
#>  7 French croissants are flaky and buttery          2
#>  8 Mexican tacos with fresh salsa                   2
#>  9 Soccer is the world's most popular sport         3
#> 10 Basketball requires speed and agility            3
#> 11 Tennis matches can last for hours                3
#> 12 Swimming is excellent cardiovascular exercise    3
```

The algorithm successfully grouped texts by topic (programming, food, sports) without any labeled training data.

```{r embedding-projection-rendered, eval = TRUE, echo = FALSE, fig.alt = "Two-dimensional embedding projection with programming, food, and sports clusters."}
projection <- embedding_cache$projection

ggplot2::ggplot(
  projection,
  ggplot2::aes(x = pc1, y = pc2, color = topic, label = label)
) +
  ggplot2::geom_point(size = 3.2, alpha = 0.9) +
  ggplot2::geom_text(nudge_y = 0.12, size = 3, show.legend = FALSE) +
  ggplot2::scale_color_manual(
    values = c(
      Programming = "#0078D4",
      Food = "#107C10",
      Sports = "#5C2D91"
    )
  ) +
  ggplot2::labs(
    title = "A two-dimensional projection makes embedding clusters visible",
    x = "Principal component 1",
    y = "Principal component 2",
    color = "Topic"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank()
  )
```

## Tips for Working with Embeddings

### Choosing a Model

Azure AI Foundry offers several embedding models:

| Model | Dimensions | Notes |
|-------|-----------|-------|
| text-embedding-ada-002 | 1,536 | Previous generation, widely used |
| text-embedding-3-small | 1,536 (configurable) | Newer, supports dimension reduction |
| text-embedding-3-large | 3,072 (configurable) | Highest quality, more expensive |

For most use cases, `text-embedding-3-small` offers a good balance of quality and cost.

### Dimension Trade-offs

Higher dimensions generally capture more nuance but:

- Require more storage space
- Take longer to compute similarities
- May not significantly improve results for simple tasks

Consider using reduced dimensions (256-512) for large-scale applications where speed matters more than precision.

### Handling Large Document Collections

For large collections, consider:

1. **Batch processing**: Embed documents in batches to avoid rate limits
2. **Caching**: Store embeddings in a database rather than regenerating them
3. **Approximate nearest neighbors**: Use libraries like `RcppAnnoy` for faster similarity search on large datasets

```{r batch-example}
# Example: Processing in batches
batch_embed <- function(texts, model, batch_size = 100) {
  n_batches <- ceiling(length(texts) / batch_size)
  results <- vector("list", n_batches)

  for (i in seq_len(n_batches)) {
    start_idx <- (i - 1) * batch_size + 1
    end_idx <- min(i * batch_size, length(texts))
    batch_texts <- texts[start_idx:end_idx]

    results[[i]] <- foundry_embed(batch_texts, model = model)

    # Brief pause to respect rate limits
    Sys.sleep(0.5)
  }

  dplyr::bind_rows(results)
}
```

### Preprocessing Text

For best results:

- Remove excessive whitespace and special characters
- Consider whether to include or exclude punctuation based on your use case
- For long documents, embed meaningful chunks (paragraphs, sentences) rather than entire documents
- Normalize text (lowercase) if case distinctions are not important for your application
