---
title: "Evaluate models and agents in Microsoft Foundry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluate models and agents in Microsoft Foundry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The Azure calls on this page replay credential-free fixtures recorded once
# with data-raw/record-doc-outputs.R and committed under vignettes/evaluations/.
# Without fixtures, only the local chunks run.
fixture_dir <- "evaluations"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
run_api <- requireNamespace("httptest2", quietly = TRUE) &&
  (recording || have_fixtures)

# Attach foundryR before start_vignette(): httptest2 only sources the package's
# inst/httptest2/start-vignette.R (which sets replay placeholders) from attached
# packages.
library(foundryR)

if (run_api) {
  httptest2::start_vignette(fixture_dir)
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_api
)
```

Microsoft Foundry can grade model and agent output for you. You describe the
test cases and the graders, Foundry runs them in your project, and the results
sit in the portal next to the deployments and agents they measured.
`foundry_evaluate()` drives that service from a data frame and returns the
grader results joined to your rows, so the analysis happens in R.

This article grades a model deployment on a labeled classification task,
compares two prompts inside one evaluation, and evaluates a prompt agent with
Foundry's agent evaluators. `vignette("evaluation-analysis")` continues from
there with intervals, paired comparisons, and checks of an LLM judge against
human labels.

If the thing you are testing is an ellmer chat running on your machine, or you
want the Inspect log viewer, use [vitals](https://vitals.tidyverse.org).
foundryR fits when the model or agent under test lives in Foundry, when you
want Foundry's built-in evaluators, or when each run should be recorded in the
project.

## Setup

Cloud evaluation uses the project endpoint and a project-scoped token. With a
project endpoint configured, every evaluation function in foundryR uses the
project route. Your account needs the Foundry User role on the project.

```{r credentials, eval = FALSE}
foundry_set_project_endpoint(
  "https://<account>.services.ai.azure.com/api/projects/<project>"
)
foundry_set_token_provider(
  foundry_token_azure_cli("https://ai.azure.com"),
  scope = "project"
)
```

The examples use one deployment to generate answers and one to act as the
judge for built-in evaluators. Replace these names with deployments in your
project. Using the same deployment for both keeps this example small; in
practice, a stronger judge that is not the model under test is the safer
choice.

```{r deployments, eval = TRUE}
target_model <- "gpt-5-nano"
judge_model <- "gpt-5-nano"
```

## Grade a model on labeled examples

Twelve course-feedback comments, each labeled with one of four themes. The
comments and labels were written for this article.

```{r feedback-data, eval = TRUE}
feedback <- tibble::tibble(
  id = 1:12,
  comment = c(
    "The lectures were clear and the slides matched what was said.",
    "Weekly quizzes were too long for the time we had.",
    "Office hours were always full, so I never got help.",
    "The textbook chapters were out of date.",
    "Feedback on the midterm came back after the final.",
    "The instructor explained regression with great examples.",
    "I could not find the practice datasets on the course site.",
    "The TA answered forum questions within a day.",
    "Grading rubrics were never posted before assignments were due.",
    "Recorded lectures had no captions.",
    "The pace was fine, but the instructor skipped the hard proofs.",
    "Tutoring sessions helped me catch up after I was sick."
  ),
  theme = c(
    "instruction", "assessment", "support", "materials",
    "assessment", "instruction", "materials", "support",
    "assessment", "materials", "instruction", "support"
  )
)
```

A string-check grader compares the generated text with the label.
`{{sample.output_text}}` is the text the target produces for a row, and
`{{item.theme}}` is the `theme` column of that row. The `ilike` operation
ignores case.

```{r theme-grader, eval = TRUE}
theme_match <- foundry_grader_string_check(
  name = "theme-match",
  input = "{{sample.output_text}}",
  reference = "{{item.theme}}",
  operation = "ilike"
)

short_prompt <- paste(
  "Classify the course feedback comment into one theme.",
  "Reply with one lowercase word: instruction, assessment, support, or materials."
)
```

With `target` set, Foundry sends each comment to the deployment, using
`short_prompt` as the developer message, and grades what comes back.
`foundry_evaluate()` checks that every `{{item.*}}` reference names a column of
`feedback` before it creates anything. It then creates an evaluation and a run
in the project and waits for the run to finish.

```{r first-run}
themes <- foundry_evaluate(
  feedback,
  graders = theme_match,
  target = target_model,
  input = "comment",
  instructions = short_prompt,
  name = "course-feedback-themes"
)
```

The result has one row per comment and grader. The columns of `feedback` come
first, followed by the grader result and the generated text.

```{r first-results}
themes |>
  dplyr::select(id, theme, .output_text, .passed)
```

```{r first-summary}
themes |>
  dplyr::summarise(
    passed = sum(.passed, na.rm = TRUE),
    cases = dplyr::n(),
    pass_rate = mean(.passed, na.rm = TRUE)
  )
```

The completed run is attached as the `"run"` attribute. Besides the pass and
fail counts, it records what Foundry measured while generating: median and
95th-percentile latency of the target and, when the deployment can be priced,
an estimated inference cost. That estimate excludes judge models and
evaluation runtime; see Microsoft's notes on [latency and estimated
cost](https://learn.microsoft.com/azure/foundry/observability/how-to/cloud-evaluation-results#review-model-target-latency-and-estimated-cost).

```{r first-run-metrics}
attr(themes, "run") |>
  dplyr::select(
    status, result_passed, result_failed,
    target_latency_p50_ms, target_latency_p95_ms,
    target_cost, target_cost_currency
  )
```

Misses are worth reading one by one. A miss can be a wrong theme or a
formatting slip such as a trailing period. The grader fails both, and a
pipeline that parses the output would fail on both too.

```{r first-misses}
themes |>
  dplyr::filter(!.passed) |>
  dplyr::select(comment, theme, .output_text)
```

## Compare two prompts in one evaluation

A new run inside the same evaluation keeps the graders fixed, and the Foundry
portal can compare runs that share an evaluation. The second prompt adds a
one-line definition of each theme.

```{r defined-prompt, eval = TRUE}
defined_prompt <- paste(
  short_prompt,
  "instruction means teaching, lectures, and explanations.",
  "assessment means quizzes, exams, grading, and feedback on work.",
  "support means office hours, tutoring, and help from course staff.",
  "materials means textbooks, slides, datasets, recordings, and the course site."
)
```

Passing `eval_id` adds a run to the evaluation created above. `wait = FALSE`
returns as soon as the run is queued, which helps when you start several runs,
for example one per deployment, and collect them afterwards.

```{r defined-run}
run_defined <- foundry_evaluate(
  feedback,
  eval_id = themes$.eval_id[[1]],
  target = target_model,
  input = "comment",
  instructions = defined_prompt,
  name = "defined-prompt",
  wait = FALSE
)

run_defined |>
  dplyr::select(run_id, status)
```

```{r defined-collect}
foundry_eval_run_wait(run_defined$eval_id, run_defined$run_id) |>
  dplyr::select(run_id, status, result_passed, result_failed)

defined <- foundry_eval_run_results(
  run_defined$eval_id,
  run_defined$run_id,
  data = feedback
)
```

Joining on `id` puts both prompts' results for each comment side by side.

```{r compare-prompts}
prompt_comparison <- dplyr::inner_join(
  dplyr::select(themes, id, theme, short_prompt = .passed),
  dplyr::select(defined, id, defined_prompt = .passed),
  by = "id"
)

dplyr::count(prompt_comparison, short_prompt, defined_prompt)
```

With twelve comments, a change of one or two cases says little.
`vignette("evaluation-analysis")` shows how to put an interval on a paired
difference and how many test cases a comparison needs.

## Evaluate an agent

Agents do things that a string match cannot score, such as following the
instructions they were given and working out what the user wants. Foundry's
agent evaluators answer those questions with a judge model. This prompt agent
is a course helpdesk with a few rules.

```{r agent-create}
helpdesk <- foundry_agent_create(
  name = "foundryr-course-helpdesk",
  model = target_model,
  instructions = paste(
    "You answer questions about the logistics of STAT 101.",
    "Problem sets are due Fridays at 5 pm. The final project is due December 12.",
    "Office hours are Tuesdays and Thursdays from 2 to 4 pm in room 210.",
    "Never change grades or discuss another student's work;",
    "refer grading questions to the instructor.",
    "Decline questions that are not about the course."
  ),
  description = "Course helpdesk used in the foundryR evaluations article."
)

helpdesk |>
  dplyr::select(agent_name, description)
```

The test cases mix questions the agent should answer, requests its rules
forbid, and questions outside its scope. The `scenario` column is only for
analysis; it travels with each item but no grader reads it.

```{r agent-cases, eval = TRUE}
helpdesk_cases <- tibble::tibble(
  scenario = c(
    "in scope", "in scope", "in scope",
    "policy", "policy",
    "out of scope", "out of scope"
  ),
  query = c(
    "When is the final project due?",
    "What time are office hours on Thursday?",
    "Where do I find this week's problem set?",
    "Can you raise my midterm grade to a B?",
    "What score did my lab partner get on the quiz?",
    "Can you recommend a laptop for gaming?",
    "Write me a cover letter for a marketing job."
  )
)
```

Task adherence judges the whole interaction against the agent's instructions,
so it reads `{{sample.output_items}}`: the agent's structured output, including
any tool calls. Intent resolution needs only the final text. Microsoft's
[agent evaluator
reference](https://learn.microsoft.com/azure/foundry/concepts/evaluation-evaluators/agent-evaluators)
lists the inputs and parameters each evaluator requires.

```{r agent-graders, eval = TRUE}
agent_graders <- list(
  foundry_grader_azure_ai(
    name = "task_adherence",
    evaluator_name = "builtin.task_adherence",
    initialization_parameters = list(deployment_name = judge_model),
    data_mapping = list(
      query = "{{item.query}}",
      response = "{{sample.output_items}}"
    )
  ),
  foundry_grader_azure_ai(
    name = "intent_resolution",
    evaluator_name = "builtin.intent_resolution",
    initialization_parameters = list(deployment_name = judge_model),
    data_mapping = list(
      query = "{{item.query}}",
      response = "{{sample.output_text}}"
    )
  )
)
```

Passing the tibble from `foundry_agent_create()` as `target` pins the run to
the version that was just created. For an existing agent, pass
`foundry_agent_reference(name, version = ...)`. Without a version, Foundry
evaluates whichever version is latest when the run starts, and foundryR prints
a reminder.

```{r agent-run}
agent_results <- foundry_evaluate(
  helpdesk_cases,
  graders = agent_graders,
  target = helpdesk,
  input = "query",
  name = "course-helpdesk-agent"
)
```

```{r agent-summary}
agent_results |>
  dplyr::group_by(scenario, .grader) |>
  dplyr::summarise(
    passed = sum(.passed, na.rm = TRUE),
    cases = dplyr::n(),
    .groups = "drop"
  )
```

The `.reason` column holds the judge's explanation. Read it before trusting a
score. A failure on an out-of-scope question can mean the agent answered when it
should have declined, or that the judge penalized a correct refusal.
`vignette("evaluation-analysis")` shows how to measure how often a judge agrees
with a person.

```{r agent-reasons}
agent_results |>
  dplyr::filter(!.passed) |>
  dplyr::select(query, .grader, .score, .reason)
```

The agent is no longer needed. The evaluations and their runs stay in the
project as a record.

```{r agent-delete}
foundry_agent_delete("foundryr-course-helpdesk") |>
  dplyr::select(agent_name, deleted)
```

## Evaluate responses your application already stored

When an application calls the Responses API with storage turned on, Foundry can
retrieve those responses by ID and grade them without replaying the requests,
which suits sampling production traffic. This path uses the lower-level
functions and is not run when this article is built.

```{r stored-responses, eval = FALSE}
prompts <- c(
  "Summarize the late-work policy for STAT 101 in one sentence.",
  "How do I ask for an extension on a problem set?"
)
answers <- purrr::map_dfr(prompts, function(prompt) {
  foundry_response(
    prompt,
    model = target_model,
    store = TRUE,
    project_endpoint = foundry_get_project_endpoint()
  )
})

stored <- foundry_eval_create(
  name = "stored-helpdesk-responses",
  data_source_config = foundry_eval_data_config(
    type = "azure_ai_source",
    scenario = "responses"
  ),
  testing_criteria = foundry_grader_azure_ai(
    name = "coherence",
    evaluator_name = "builtin.coherence",
    initialization_parameters = list(model = judge_model)
  )
)

stored_run <- foundry_eval_run_create(
  stored$eval_id,
  data_source = foundry_eval_run_data(response_ids = answers$response_id)
)
foundry_eval_run_wait(stored$eval_id, stored_run$run_id)
foundry_eval_run_output_items(stored$eval_id, stored_run$run_id)
```

Foundry reads the query and the answer from each stored response, so these
graders need no data mapping. Stored-response runs accept response IDs only
inline, which is how `foundry_eval_run_data(response_ids = ...)` sends them.

## Safety evaluators

Foundry's risk and safety evaluators, such as violence, self-harm, hate and
unfairness, and indirect attack, run on Microsoft-managed models and are
available only in [some
regions](https://learn.microsoft.com/azure/foundry/concepts/evaluation-evaluators/risk-safety-evaluators).
Add them like any other built-in evaluator. They are left out of the runs above
so that this article can be rebuilt in any region.

```{r safety-grader, eval = FALSE}
foundry_grader_azure_ai(
  name = "violence",
  evaluator_name = "builtin.violence",
  data_mapping = list(
    query = "{{item.query}}",
    response = "{{sample.output_text}}"
  )
)
```

## Finding the results later

Every evaluation above is still in the project. `foundry_evals()`,
`foundry_eval_runs()`, and `foundry_eval_run_output_items()` list them from R,
and each run has a report in the Foundry portal. To rebuild a joined tibble,
pass the same data frame to `foundry_eval_run_results()`. foundryR compares the
item fields that Foundry echoes with your rows and stops if they differ, so a
reordered or edited data frame cannot be matched to the wrong results.

```{r cleanup, include = FALSE}
if (run_api) {
  httptest2::end_vignette()
}
```
