---
title: "Analyze evaluation results with uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyze evaluation results with uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
have_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
points <- function(x) sprintf("%.0f", 100 * x)
```

```{r library, message = FALSE}
library(foundryR)
library(dplyr)
```

An evaluation run reports a pass rate, and a pass rate computed from a hundred
test cases carries sampling error. Two prompts that score 71% and 78% on the
same test set may or may not differ in a way that holds up on new inputs. This
article works through four checks that help separate a real difference from
noise: intervals for pass rates, a paired comparison of two targets on the same
cases, a sample-size calculation, and a test of the LLM judge against human
labels.

The examples use simulated results shaped like the output of
`foundry_evaluate()`, so the article runs without credentials. None of the
numbers below come from a model. `vignette("evaluations")` shows how to produce
real results.

## Simulated results

The simulation scores 120 support tickets in three segments with two targets,
`model-a` and `model-b`, and one grader. Some tickets are harder than others for
both targets, which is what makes the two results for the same ticket
correlated. By construction, `model-b` is stronger overall and more so on
technical tickets. Because the simulation knows each ticket's pass probability,
it also gives the expected pass rates that the observed results estimate.

```{r simulate}
set.seed(20260927)
n_cases <- 120
tickets <- tibble(
  case_id = seq_len(n_cases),
  segment = rep(c("billing", "technical", "account"), times = c(50, 40, 30)),
  difficulty = rnorm(n_cases)
)

simulate_run <- function(tickets, skill, technical_bonus = 0) {
  shift <- c(billing = 0.6, technical = -0.5, account = 0.2)[tickets$segment]
  bonus <- ifelse(tickets$segment == "technical", technical_bonus, 0)
  p_pass <- plogis(skill - 1.2 * tickets$difficulty + shift + bonus)
  tickets |>
    mutate(.grader = "label-match", .passed = runif(n()) < p_pass, p_true = p_pass) |>
    select(-difficulty)
}

simulated <- bind_rows(
  "model-a" = simulate_run(tickets, skill = 1.0),
  "model-b" = simulate_run(tickets, skill = 1.3, technical_bonus = 0.5),
  .id = "target"
)
expected <- simulated |>
  group_by(target, segment) |>
  summarise(expected_pass_rate = mean(p_true), .groups = "drop")
results <- select(simulated, -p_true)
results
```

With real runs, `results_a` and `results_b` would come from two
`foundry_evaluate()` calls on the same data frame, and
`bind_rows("model-a" = results_a, "model-b" = results_b, .id = "target")` would
stack them the same way.

## Pass rates with intervals

The Wilson score interval stays inside 0 and 1 and behaves well near the edges,
where the textbook normal interval does not.

```{r pass-rates}
wilson_interval <- function(passed, n, level = 0.95) {
  z <- qnorm(1 - (1 - level) / 2)
  p <- passed / n
  center <- (p + z^2 / (2 * n)) / (1 + z^2 / n)
  half <- z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2)) / (1 + z^2 / n)
  tibble(lower = center - half, upper = center + half)
}

pass_rates <- results |>
  group_by(target) |>
  summarise(passed = sum(.passed), cases = n(), .groups = "drop") |>
  mutate(pass_rate = passed / cases, wilson_interval(passed, cases))

pass_rates
```

With `r n_cases` cases, each interval reaches about
`r points(mean(pass_rates$upper - pass_rates$lower) / 2)` percentage points on
either side of its estimate. Differences smaller than that deserve the checks
below before anyone acts on them.

## Compare the targets case by case

Both targets answered the same tickets, so compare them ticket by ticket.
Tickets where both passed, or both failed, say nothing about which target is
better. Only the discordant tickets do.

```{r paired}
paired <- inner_join(
  results |> filter(target == "model-a") |> select(case_id, segment, a = .passed),
  results |> filter(target == "model-b") |> select(case_id, b = .passed),
  by = "case_id"
)
count(paired, a, b)

only_b <- sum(!paired$a & paired$b)
only_a <- sum(paired$a & !paired$b)
mcnemar <- binom.test(only_b, only_b + only_a)
mcnemar$p.value
```

`model-b` passed `r sum(paired$b)` tickets and `model-a` passed
`r sum(paired$a)`. Of the `r only_a + only_b` discordant tickets, `model-b`
passed `r only_b` that `model-a` failed, and `model-a` passed `r only_a` that
`model-b` failed. McNemar's exact test on those tickets gives
p = `r format.pval(mcnemar$p.value, digits = 2)`.

A bootstrap that resamples tickets, not rows, gives an interval for the
difference in pass rates. Resample at the level where the test cases were
drawn. If one conversation contributes several items, resample conversations.

```{r paired-interval}
difference <- mean(paired$b) - mean(paired$a)
bootstrap <- replicate(2000, {
  rows <- sample.int(nrow(paired), replace = TRUE)
  mean(paired$b[rows]) - mean(paired$a[rows])
})
paired_ci <- quantile(bootstrap, c(0.025, 0.975), names = FALSE)
unpaired_ci <- prop.test(
  c(sum(paired$b), sum(paired$a)),
  c(nrow(paired), nrow(paired)),
  correct = FALSE
)$conf.int

tibble(
  method = c("paired bootstrap over tickets", "unpaired two-sample interval"),
  difference = difference,
  lower = c(paired_ci[[1]], unpaired_ci[[1]]),
  upper = c(paired_ci[[2]], unpaired_ci[[2]])
)
```

```{r true-difference, include = FALSE}
expected_overall <- simulated |>
  group_by(target) |>
  summarise(expected_pass_rate = mean(p_true), .groups = "drop")
true_difference <- diff(expected_overall$expected_pass_rate[match(c("model-a", "model-b"), expected_overall$target)])
```

The two results for a ticket are correlated
(r = `r sprintf("%.2f", cor(paired$a, paired$b))`), so the paired interval is
`r if (diff(paired_ci) < diff(unpaired_ci)) "narrower" else "no narrower"` than
the unpaired one, which treats the runs as independent samples. The observed
difference of `r points(difference)` points has a paired 95% interval from
`r points(paired_ci[[1]])` to `r points(paired_ci[[2]])` points, which
`r if (paired_ci[[1]] <= 0 && paired_ci[[2]] >= 0) "includes zero" else "excludes zero"`.
The simulation's expected pass rates differ by `r points(true_difference)`
points, so `model-b` really is better here. A test set of `r n_cases` tickets is
too small to show it reliably.

The quantity being estimated is the difference in pass rate between these two
targets on tickets like these, with these prompts, sampling settings, and
grader. It says little about inputs unlike the test set, and a test set you
have tuned prompts against will overstate how well the winning prompt
generalizes. Keep a held-out set for the final comparison.

## Where the targets fail

Segment pass rates show where to look next, but each segment has fewer cases, so
its interval is wider. Scanning many segments for the largest gap also finds
gaps that are only noise.

```{r by-segment}
by_segment <- results |>
  group_by(target, segment) |>
  summarise(passed = sum(.passed), cases = n(), .groups = "drop") |>
  mutate(pass_rate = passed / cases, wilson_interval(passed, cases))

by_segment
```

```{r segment-reversals, include = FALSE}
segment_gaps <- inner_join(by_segment, expected, by = c("target", "segment")) |>
  group_by(segment) |>
  summarise(
    observed_gap = pass_rate[target == "model-b"] - pass_rate[target == "model-a"],
    expected_gap = expected_pass_rate[target == "model-b"] - expected_pass_rate[target == "model-a"],
    .groups = "drop"
  )
reversed <- segment_gaps[sign(segment_gaps$observed_gap) != sign(segment_gaps$expected_gap), ]
reversal_sentence <- if (nrow(reversed) > 0) {
  sprintf(
    paste(
      "The %s segment shows how easily that happens. By construction `model-b` is",
      "better on %s tickets, by %s points in expectation, yet this sample shows it",
      "%s points lower."
    ),
    reversed$segment[[1]],
    reversed$segment[[1]],
    points(reversed$expected_gap[[1]]),
    points(abs(reversed$observed_gap[[1]]))
  )
} else {
  paste(
    "Here every observed segment gap points the same way as the expected one,",
    "but the intervals are wide enough that a different sample could reverse some of them."
  )
}
```

`r reversal_sentence`

```{r segment-chart-data, include = FALSE, eval = have_ggplot2}
segment_levels <- c("account", "technical", "billing", "All tickets")
chart_data <- bind_rows(
  mutate(pass_rates, segment = "All tickets"),
  by_segment
) |>
  mutate(
    segment = factor(segment, levels = segment_levels),
    y = as.integer(segment) + ifelse(target == "model-a", 0.16, -0.16)
  )
stopifnot(!anyNA(chart_data$segment))

series_colors <- c("model-a" = "#065D92", "model-b" = "#AD4D0E")
```

```{r segment-chart-title, include = FALSE, eval = have_ggplot2}
seg_wide <- inner_join(
  filter(by_segment, target == "model-a") |> select(segment, a_lower = lower, a_upper = upper),
  filter(by_segment, target == "model-b") |> select(segment, b_lower = lower, b_upper = upper),
  by = "segment"
)
separated <- seg_wide$segment[seg_wide$b_lower > seg_wide$a_upper | seg_wide$a_lower > seg_wide$b_upper]
gap_cases <- sum(paired$b) - sum(paired$a)
chart_title <- sprintf(
  "On %d tickets, model-b passed %d %s than model-a; %s",
  n_cases,
  abs(gap_cases),
  if (gap_cases >= 0) "more" else "fewer",
  if (length(separated) == 0L) {
    "every segment's intervals overlap"
  } else {
    paste("the intervals separate only for", paste(separated, collapse = " and "))
  }
)
describe_row <- function(label) {
  rows <- chart_data[chart_data$segment == label, ]
  paste(
    sprintf("%s %s%% (%s to %s)", rows$target, points(rows$pass_rate), points(rows$lower), points(rows$upper)),
    collapse = ", "
  )
}
chart_alt <- paste0(
  chart_title, ". Dot and interval chart of pass rates with 95% Wilson intervals. ",
  paste(
    vapply(rev(segment_levels), function(label) paste0(label, ": ", describe_row(label)), character(1)),
    collapse = "; "
  ),
  "."
)
x_floor <- floor(min(chart_data$lower) * 10) / 10
```

```{r segment-chart, echo = FALSE, eval = have_ggplot2, fig.width = 7, fig.height = 3.9, out.width = "100%", fig.alt = get0("chart_alt", ifnotfound = "Pass rates with 95% intervals by segment for two targets.")}
top_row <- chart_data[chart_data$segment == "All tickets", ]
wrap_title <- function(text, width) paste(strwrap(text, width), collapse = "\n")

ggplot2::ggplot(chart_data, ggplot2::aes(y = y, colour = target)) +
  ggplot2::geom_segment(
    ggplot2::aes(x = lower, xend = upper, yend = y),
    linewidth = 0.9
  ) +
  ggplot2::geom_point(ggplot2::aes(x = pass_rate, shape = target), size = 2.8) +
  ggplot2::geom_text(
    data = top_row,
    ggplot2::aes(x = upper, label = target),
    hjust = -0.15,
    size = 3.6
  ) +
  ggplot2::scale_colour_manual(values = series_colors, guide = "none") +
  ggplot2::scale_shape_manual(values = c("model-a" = 16, "model-b" = 17), guide = "none") +
  ggplot2::scale_x_continuous(
    breaks = seq(x_floor, 1, by = 0.1),
    labels = function(x) paste0(round(100 * x), "%"),
    expand = ggplot2::expansion(mult = c(0.02, 0.02))
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq_along(segment_levels),
    labels = segment_levels,
    expand = ggplot2::expansion(add = 0.45)
  ) +
  ggplot2::coord_cartesian(xlim = c(x_floor, 1.08), clip = "off") +
  ggplot2::labs(
    title = wrap_title(chart_title, 70),
    subtitle = "Pass rate with 95% Wilson interval, by ticket segment",
    x = "Pass rate",
    y = NULL,
    caption = wrap_title(
      paste(
        "Simulated results for 120 tickets. Segment intervals use only that segment's tickets;",
        "comparing many segments invites false positives."
      ),
      105
    )
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    plot.title.position = "plot",
    plot.subtitle = ggplot2::element_text(color = "grey30"),
    plot.caption = ggplot2::element_text(color = "grey40", hjust = 0),
    plot.caption.position = "plot",
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_text(size = 11, color = "grey15"),
    plot.margin = ggplot2::margin(8, 40, 8, 8)
  )
```

## How many test cases a comparison needs

The interval half-width for a single pass rate shrinks with the square root of
the number of cases. Near a pass rate of 80%:

```{r margins}
margin_of_error <- function(p, n) qnorm(0.975) * sqrt(p * (1 - p) / n)

tibble(cases = c(25, 50, 100, 200, 400, 800)) |>
  mutate(plus_or_minus_points = round(100 * margin_of_error(0.8, cases), 1))
```

For a paired comparison, what matters is how often the two targets disagree.
The normal approximation for McNemar's test (Connor, 1987) gives the number of
cases needed to detect a difference `d` with 80% power when a share `psi` of
cases is discordant.

```{r paired-sample-size}
paired_cases_needed <- function(d, psi, alpha = 0.05, power = 0.8) {
  z_alpha <- qnorm(1 - alpha / 2)
  z_beta <- qnorm(power)
  ceiling((z_alpha * sqrt(psi) + z_beta * sqrt(psi - d^2))^2 / d^2)
}

observed_psi <- (only_a + only_b) / nrow(paired)
tibble(d = c(0.03, 0.05, 0.10)) |>
  filter(d <= observed_psi) |>
  mutate(cases_needed = paired_cases_needed(d, psi = observed_psi))
```

At the discordance observed here (`r points(observed_psi)`% of tickets),
detecting a 5-point improvement with 80% power takes about
`r format(paired_cases_needed(0.05, observed_psi), big.mark = ",")` tickets,
well beyond the `r n_cases` in this test set.

## Check the judge before trusting it

When the grader is a model, its errors flow into every pass rate it reports.
Before comparing targets with an LLM judge, have people label a random sample
of graded responses without seeing the judge's verdict, then measure agreement.
In the simulated validation sample of 80 responses below, the judge passes 92%
of the responses people pass and 30% of the responses people fail.

```{r judge-validation}
validation <- tibble(human_pass = runif(80) < 0.7) |>
  mutate(
    judge_pass = ifelse(human_pass, runif(n()) < 0.92, runif(n()) < 0.30),
    human = ifelse(human_pass, "pass", "fail"),
    judge = ifelse(judge_pass, "pass", "fail")
  )

foundry_agreement(validation, estimate = "judge", truth = "human")
count(validation, human, judge)
```

```{r judge-counts, include = FALSE}
false_pass <- sum(validation$judge == "pass" & validation$human == "fail")
false_fail <- sum(validation$judge == "fail" & validation$human == "pass")
judge_rate <- points(mean(validation$judge == "pass"))
human_rate <- points(mean(validation$human == "pass"))
judge_bias <- if (false_pass == false_fail) {
  sprintf(
    paste(
      "The errors cancel in the aggregate, and the judge and people both report a",
      "%s%% pass rate. Matching pass rates are not evidence of a good judge.",
      "The two kinds of error can stop cancelling when the mix of responses",
      "changes, for example between two targets, and then the judge distorts the",
      "comparison."
    ),
    judge_rate
  )
} else {
  sprintf(
    "The judge reports a %s%% pass rate against %s%% from people, so it makes targets look %s than they are.",
    judge_rate,
    human_rate,
    if (false_pass > false_fail) "better" else "worse"
  )
}
```

Cohen's kappa discounts the agreement two raters would reach by chance, so it
is lower than raw accuracy. The judge disagrees with people on
`r false_pass + false_fail` of 80 responses: `r false_pass` it passed and people
failed, and `r false_fail` it failed and people passed. `r judge_bias`

If agreement is low, fix the evaluator before comparing runs: tighten the rubric
of a label or score grader, adjust its threshold, or use a stronger judge
model. To estimate a pass rate that accounts for judge error, use methods built
for inference on predicted labels, such as those in the
[ipd](https://CRAN.R-project.org/package=ipd) package, rather than an ad hoc
correction.

## Weigh quality against cost and latency

For model-target runs, the run tibble from `foundry_eval_run_get()` or
`attr(results, "run")` carries Foundry's measurements of target latency and
estimated cost. The figures below are invented for illustration.

```{r cost-latency}
runs <- tibble(
  target = c("model-a", "model-b"),
  target_latency_p50_ms = c(640, 1420),
  target_latency_p95_ms = c(1900, 4100),
  target_cost = c(0.018, 0.071),
  target_cost_currency = "USD"
)

runs |>
  left_join(select(pass_rates, target, passed, cases), by = "target") |>
  mutate(cost_per_1000_cases = 1000 * target_cost / cases) |>
  select(target, target_latency_p95_ms, target_cost, passed, cost_per_1000_cases)
```

```{r cost-increment, include = FALSE}
extra_passes <- diff(pass_rates$passed[match(c("model-a", "model-b"), pass_rates$target)])
extra_cost <- diff(runs$target_cost)
```

In this example, `model-b` costs about `r sprintf("%.1f", runs$target_cost[2] / runs$target_cost[1])`
times as much per ticket and its 95th-percentile latency is more than twice as
long. It passed `r extra_passes` more tickets, which works out to
`r if (extra_passes > 0) sprintf("$%.4f", extra_cost / extra_passes) else "no gain"`
per additional pass on this test set. That figure inherits the uncertainty of
the pass-rate difference. With an interval that
`r if (paired_ci[[1]] <= 0 && paired_ci[[2]] >= 0) "includes zero, the true cost per additional pass could be far higher, or there may be no gain at all" else "excludes zero, the direction is clear but the size is not"`.
Foundry's cost estimate covers target inference only; judge models and
evaluation runtime are billed separately.

## Report the method

When you report an evaluation, give the number of test cases, the interval
method, how paired runs were compared, and how the judge was validated. The
Foundry portal's run comparison computes its own statistics, which can differ
from the paired analysis here, so say which one a reported number comes from.

Connor, R. J. (1987). Sample size for testing differences in proportions for the
paired-sample design. *Biometrics*, 43(1), 207-211.
