---
title: "Files and Batch API Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Files and Batch API Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(foundryR)
```

The Batch API is useful when your research task has hundreds or thousands of
independent rows: survey coding, abstract screening, entity extraction, document
classification, or large-scale summarization.

## Prepare JSONL locally

`foundry_batch_requests()` runs locally. It converts a data frame into the JSON
Lines shape expected by the Batch API.

```{r batch-jsonl}
survey <- data.frame(
  id = c("resp-001", "resp-002", "resp-003"),
  response = c(
    "The workshop was clear and practical.",
    "I liked the examples but wanted more time.",
    "The setup instructions were confusing."
  )
)

jsonl <- tempfile(fileext = ".jsonl")

request_file <- foundry_batch_requests(
  survey,
  input = "response",
  path = jsonl,
  model = "gpt-4.1",
  custom_id = "id",
  body = list(
    instructions = "Classify the response sentiment as positive, neutral, or negative."
  )
)

request_file
head(readLines(jsonl), 2)
```

## Upload and create a batch

Uploading and batch creation call the Foundry service, so these chunks are not
run while building the vignette.

```{r upload-create, eval = FALSE}
file <- foundry_file_upload(jsonl, purpose = "batch")

batch <- foundry_batch_create(
  input_file_id = file$file_id,
  endpoint = "/v1/responses"
)
```

The returned objects are tidy tibbles:

```{r upload-create-output}
file <- tibble::tibble(
  file_id = "file_abc123",
  filename = basename(jsonl),
  purpose = "batch",
  status = "processed",
  bytes = file.size(jsonl),
  created_at = as.POSIXct("2026-06-24 12:00:00", tz = "UTC")
)

batch <- tibble::tibble(
  batch_id = "batch_abc123",
  status = "validating",
  endpoint = "/v1/responses",
  input_file_id = file$file_id,
  output_file_id = NA_character_,
  error_file_id = NA_character_,
  completion_window = "24h",
  request_counts_total = 3L,
  request_counts_completed = 0L,
  request_counts_failed = 0L
)

file
batch
```

## Poll and download results

```{r poll-download, eval = FALSE}
foundry_batch_get(batch$batch_id)

foundry_file_download(
  file_id = batch$output_file_id,
  path = "batch-output.jsonl"
)
```

Output files are JSONL too. Read a few lines first before parsing a large job:

```{r output-shape}
output_lines <- c(
  '{"custom_id":"resp-001","response":{"body":{"output_text":"positive"}}}',
  '{"custom_id":"resp-002","response":{"body":{"output_text":"neutral"}}}'
)

head(output_lines)
```

## Practical advice

- Start with 10 to 20 rows and inspect the output before scaling up.
- Use stable `custom_id` values so results join back to your data frame.
- Store prompts and schema versions with your analysis code for reproducibility.
- Download both output and error files when a batch finishes.
