---
title: "foundryR vs ellmer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{foundryR vs ellmer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The Azure-backed example on this page runs against recorded, credential-free
# fixtures (recorded once with data-raw/record-doc-outputs.R and committed under
# vignettes/foundryr-vs-ellmer/). The ellmer -> foundryR schema conversion is a
# pure-local call and always runs when ellmer is installed, so its real output is
# shown everywhere. Nothing here is fabricated.
fixture_dir <- "foundryr-vs-ellmer"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
run_api <- recording || have_fixtures
have_ellmer <- requireNamespace("ellmer", quietly = TRUE)

if (run_api) {
  httptest2::start_vignette(fixture_dir)
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_api
)
```

```{r library, eval = TRUE}
library(foundryR)
```

foundryR and ellmer both make language-model work possible from R. They are not
substitutes for every use case. ellmer is the better fit for provider-portable
chat. foundryR is the better fit when the work is tied to Azure AI Foundry and
the output needs to become data. The two are complementary: you can describe a
structure once with ellmer's type system and hand it to foundryR for strict,
tibble-shaped extraction.

## Comparison

| Question | Use foundryR | Use ellmer |
| --- | --- | --- |
| Are you committed to Azure AI Foundry? | Yes | Sometimes |
| Do you need Azure AI Content Safety? | Yes | No |
| Do you need Azure's Files and Batch APIs? | Yes | No |
| Do you need strict schema-constrained extraction as tibbles? | Yes | Sometimes |
| Do you need embeddings inside a dataframe workflow? | Yes | Sometimes |
| Do you need `step_foundry_embed()` in a tidymodels recipe? | Yes | No |
| Do you need multi-provider chat? | No | Yes |
| Do you need interactive streaming chat? | No | Yes |
| Do you need a chat-first tool-calling agent interface? | Basic Responses API loop | Yes |

## Where foundryR fits

Use foundryR when your work starts or ends in a dataframe. Common examples are:

- Coding open-ended survey responses with a strict JSON Schema.
- Running low-cost batch annotation jobs through Azure's Batch API.
- Embedding documents for clustering, semantic search, or near-duplicate checks.
- Adding embeddings to tidymodels recipes.
- Moderating user text, detecting prompt injection, or checking groundedness.
- Keeping model metadata, citations, tool calls, and token use in tibble columns.

Azure's own Batch API matters here. ellmer supports batch workflows for native
OpenAI and Anthropic endpoints, but it does not target Azure's Files and Batch
API surface.

## Where ellmer fits

Use ellmer when chat is the product interface. It gives you a provider-neutral
chat abstraction, interactive streaming, and a mature chat-first tool-calling
workflow. That is the right shape for assistants, notebooks where you want token
streaming, or applications that may move between providers.

foundryR deliberately does not implement streaming. For streaming chat in R, use
ellmer.

## Interop: reuse an ellmer type as a foundryR schema

If you already describe your data with ellmer's type system, you do not have to
rewrite it. `as_foundry_schema()` converts an `ellmer::type_object()` into the
strict JSON Schema that `foundry_extract()` expects. This conversion is a local
operation -- no network, no credentials -- so its output is shown here directly.

```{r ellmer-interop, eval = have_ellmer}
library(ellmer)

# Describe the structure you want with ellmer's type system.
sentiment_spec <- type_object(
  sentiment = type_enum(
    c("positive", "negative", "neutral"),
    description = "Overall sentiment of the response."
  ),
  theme = type_string("A short theme label for the response.")
)

# Convert it into a foundryR schema for strict extraction.
sentiment_schema <- as_foundry_schema(sentiment_spec)
str(sentiment_schema)
```

## Extraction with foundryR

Hand the converted schema to `foundry_extract()`. foundryR sends each input
through the Responses API with strict decoding and returns one tidy row per
input -- the model's output has already become data.

```{r extraction-example, eval = run_api && have_ellmer}
foundry_extract(
  c("The lesson was clear.", "I wanted more examples."),
  schema = sentiment_schema,
  model = "gpt-5.5"
)
```

You can also build the same schema natively with foundryR's `foundry_schema()`
and `schema_*()` constructors when you would rather not depend on ellmer. See
`vignette("responses-api")` for the native path, and `vignette("content-safety")`
for using Azure Content Safety as a pipeline gate.

The practical rule is simple: use ellmer when you need the best R chat client,
and use foundryR when you need Azure AI Foundry results as data -- and use
`as_foundry_schema()` when you want both.

```{r cleanup, include = FALSE}
if (run_api) {
  httptest2::end_vignette()
}
```
