---
title: "foundryR vs ellmer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{foundryR vs ellmer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

foundryR and ellmer both make language-model work possible from R. They are not
substitutes for every use case. ellmer is the better fit for provider-portable
chat. foundryR is the better fit when the work is tied to Azure AI Foundry and
the output needs to become data.

## Comparison

| Question | Use foundryR | Use ellmer |
| --- | --- | --- |
| Are you committed to Azure AI Foundry? | Yes | Sometimes |
| Do you need Azure AI Content Safety? | Yes | No |
| Do you need Azure's Files and Batch APIs? | Yes | No |
| Do you need strict schema-constrained extraction as tibbles? | Yes | Sometimes |
| Do you need embeddings inside a dataframe workflow? | Yes | Sometimes |
| Do you need `step_foundry_embed()` in a tidymodels recipe? | Yes | No |
| Do you need multi-provider chat? | No | Yes |
| Do you need interactive streaming chat? | No | Yes |
| Do you need a chat-first tool-calling agent interface? | Basic Responses API loop | Yes |

## Where foundryR fits

Use foundryR when your work starts or ends in a dataframe. Common examples are:

- Coding open-ended survey responses with a strict JSON Schema.
- Running low-cost batch annotation jobs through Azure's Batch API.
- Embedding documents for clustering, semantic search, or near-duplicate checks.
- Adding embeddings to tidymodels recipes.
- Moderating user text, detecting prompt injection, or checking groundedness.
- Keeping model metadata, citations, tool calls, and token use in tibble columns.

Azure's own Batch API matters here. ellmer supports batch workflows for native
OpenAI and Anthropic endpoints, but it does not target Azure's Files and Batch
API surface.

## Where ellmer fits

Use ellmer when chat is the product interface. It gives you a provider-neutral
chat abstraction, interactive streaming, and a mature chat-first tool-calling
workflow. That is the right shape for assistants, notebooks where you want token
streaming, or applications that may move between providers.

foundryR deliberately does not implement streaming. For streaming chat in R, use
ellmer.

## Example: strict extraction with foundryR

```{r extraction-example}
library(foundryR)

schema <- list(
  type = "object",
  properties = list(
    sentiment = list(type = "string", enum = c("positive", "negative", "neutral")),
    theme = list(type = "string")
  ),
  required = c("sentiment", "theme"),
  additionalProperties = FALSE
)

foundry_extract(
  c("The lesson was clear.", "I wanted more examples."),
  schema = schema,
  model = "my-gpt4"
)
```

## Example: Azure Content Safety as a pipeline gate

```{r safety-example}
answer <- "The study enrolled 120 participants."
source <- "The study enrolled 120 participants."

foundry_groundedness(
  text = answer,
  grounding_sources = source,
  query = "How many participants were enrolled?",
  task = "QnA"
)
```

The practical rule is simple: use ellmer when you need the best R chat client,
and use foundryR when you need Azure AI Foundry results as data.
