---
title: "Image Generation with DALL-E"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Image Generation with DALL-E}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

foundryR integrates with DALL-E models deployed on Azure AI Foundry to generate images from text descriptions. Whether you need visualizations for reports, creative assets, or synthetic data, `foundry_image()` provides a tidy interface to AI image generation.

## Prerequisites

### Deploying a DALL-E Model

Before using image generation, deploy a DALL-E model in Azure AI Foundry:

1. Go to [Azure AI Foundry](https://ai.azure.com) or the Azure Portal
2. Navigate to your Azure OpenAI resource, then **Model deployments**
3. Click **+ Deploy model**, then **Deploy base model**
4. Select **dall-e-3** (recommended) or **dall-e-2**
5. Give it a deployment name (e.g., `my-dalle`)

### Configuration

DALL-E may be deployed on a different Azure resource than your chat models. foundryR supports separate configuration for image generation:

```{r config}
library(foundryR)

# If DALL-E is on the same resource as your chat models:
foundry_set_endpoint(Sys.getenv("AZURE_FOUNDRY_ENDPOINT"))
foundry_set_key("your-api-key")

# If DALL-E is on a different resource:
foundry_set_image_endpoint(Sys.getenv("AZURE_FOUNDRY_IMAGE_ENDPOINT"))
foundry_set_image_key("your-dalle-api-key")  # Only if different from main key

# Set default DALL-E deployment
Sys.setenv(AZURE_FOUNDRY_IMAGE_MODEL = "my-dalle")
```

## Generating Images with foundry_image()

### Basic Usage

Generate an image with a simple text prompt:

```{r basic}
library(foundryR)

result <- foundry_image(
  "A serene mountain landscape at sunset with a reflective lake",
  model = "my-dalle"
)

result
#> # A tibble: 1 × 5
#>   prompt                                     revised_prompt               url                    b64_json created
#>   <chr>                                      <chr>                        <chr>                  <chr>    <dttm>
#> 1 A serene mountain landscape at sunset...   A breathtaking mountain...   https://dalleprodsec...NA       2024-01-15 10:30:00
```
The result includes:

- **prompt**: Your original prompt
- **revised_prompt**: DALL-E 3's enhanced interpretation (more detailed)
- **url**: Temporary URL to the generated image
- **b64_json**: Base64-encoded image data (if requested)
- **created**: Timestamp of generation

### Viewing the Image

Open the generated image in your browser:

```{r view}
# View in browser
browseURL(result$url)

# Or in RStudio Viewer (if you have the right setup)
# htmltools::browsable(htmltools::img(src = result$url))
```

**Important**: Image URLs are temporary and expire after a short time. Save images locally for permanent storage.

### Saving Images

Use `foundry_save_image()` to download and save images:

```{r save}
result <- foundry_image(
  "A futuristic cityscape with flying cars",
  model = "my-dalle"
)

# Save to file
foundry_save_image(result, "cityscape.png")
#> ✓ Image saved to cityscape.png (from URL)
```

## Customizing Image Generation

### Image Size

DALL-E supports different image sizes:

```{r sizes}
# DALL-E 3 sizes
# - 1024x1024 (square, default)
# - 1792x1024 (landscape)
# - 1024x1792 (portrait)

# Landscape image
result <- foundry_image(
  "A panoramic view of the Grand Canyon",
  model = "my-dalle",
  size = "1792x1024"
)

# Portrait image
result <- foundry_image(
  "A tall redwood tree reaching toward the sky",
  model = "my-dalle",
  size = "1024x1792"
)
```

### Quality Settings

DALL-E 3 offers standard and HD quality:

```{r quality}
# Standard quality (faster, less detailed)
result <- foundry_image(
  "A simple icon of a house",
  model = "my-dalle",
  quality = "standard"
)

# HD quality (slower, finer details)
result <- foundry_image(
  "An intricate mandala pattern with fine details",
  model = "my-dalle",
  quality = "hd"
)
```

### Style Options

Control the artistic style:

```{r style}
# Vivid style (hyper-real, dramatic)
result <- foundry_image(
  "A dragon breathing fire",
  model = "my-dalle",
  style = "vivid"
)

# Natural style (more realistic, less dramatic)
result <- foundry_image(
  "A golden retriever in a park",
  model = "my-dalle",
  style = "natural"
)
```

### Multiple Images

Generate multiple variations at once:

```{r multiple}
# Generate 3 variations
results <- foundry_image(
  "An abstract representation of data science",
  model = "my-dalle",
  n = 3
)

results
#> # A tibble: 3 × 5
#>   prompt                                   revised_prompt  url       b64_json created
#>   <chr>                                    <chr>           <chr>     <chr>    <dttm>
#> 1 An abstract representation of data...   A creative...   https://… NA       2024-01-15 10:30:00
#> 2 An abstract representation of data...   An artistic...  https://… NA       2024-01-15 10:30:00
#> 3 An abstract representation of data...   A vibrant...    https://… NA       2024-01-15 10:30:00

# Save each image
for (i in seq_len(nrow(results))) {
  foundry_save_image(results, paste0("abstract_", i, ".png"), index = i)
}
```

### Base64 Response Format

For programmatic use, request base64-encoded images instead of URLs:

```{r base64}
result <- foundry_image(
  "A simple logo design",
  model = "my-dalle",
  response_format = "b64_json"
)

# The image data is in the b64_json column
nchar(result$b64_json)  # Large base64 string
#> [1] 1234567

# Save from base64
foundry_save_image(result, "logo.png")
#> ✓ Image saved to logo.png (from base64)
```

Base64 is useful when you need to:
- Store images in databases
- Embed images directly in HTML/documents
- Process images without downloading from URLs

## Prompt Engineering Tips

### Be Specific

Vague prompts produce unpredictable results. Add details about:

- **Subject**: What is the main focus?
- **Style**: Photorealistic, watercolor, digital art, sketch?
- **Composition**: Close-up, wide angle, aerial view?
- **Lighting**: Golden hour, dramatic shadows, soft diffused light?
- **Mood**: Serene, energetic, mysterious?

```{r specific}
# Vague prompt
foundry_image("a cat", model = "my-dalle")

# Specific prompt
foundry_image(
  "A fluffy orange tabby cat sitting on a windowsill,
   soft morning light streaming through lace curtains,
   photorealistic style, shallow depth of field",
  model = "my-dalle"
)
```

### Use DALL-E 3's Revised Prompts

DALL-E 3 automatically enhances your prompts. Check `revised_prompt` to learn what works:

```{r revised}
result <- foundry_image("sunset", model = "my-dalle")

# See how DALL-E interpreted your prompt
cat(result$revised_prompt)
#> A breathtaking sunset over a calm ocean, with vibrant hues of
#> orange, pink, and purple reflecting off the water's surface...
```

### Iterative Refinement

Use the revised prompt as inspiration for your next attempt:

```{r iterate}
# First attempt
r1 <- foundry_image("a robot", model = "my-dalle")

# Refine based on what you liked/disliked
r2 <- foundry_image(
  "A friendly humanoid robot with soft rounded edges,
   warm LED eyes, and a brushed aluminum finish,
   standing in a modern home environment",
  model = "my-dalle"
)
```

## Use Cases

### Data Visualization Enhancement

Generate custom illustrations for reports:

```{r dataviz}
# Create a conceptual illustration for a report section
foundry_image(
  "A clean, professional infographic-style illustration showing
   data flowing from multiple sources into a central analytics hub,
   using a blue and white color scheme, minimalist corporate style",
  model = "my-dalle",
  size = "1792x1024"
) |>
  foundry_save_image("data_flow_illustration.png")
```

### Placeholder Images for Prototypes

Quickly generate placeholder images for UI mockups:

```{r placeholder}
categories <- c("technology", "nature", "food", "travel")

for (cat in categories) {
  foundry_image(
    paste("A beautiful photograph representing", cat,
          "suitable for a website hero image"),
    model = "my-dalle"
  ) |>
    foundry_save_image(paste0("hero_", cat, ".png"))
}
```

### Synthetic Training Data

Generate diverse images for ML training (with appropriate caution):

```{r synthetic}
# Generate variations of a product for training an object detector
variations <- c(
  "on a white background, studio lighting",
  "on a wooden table, natural lighting",
  "held in a person's hand, outdoor setting",
  "on a store shelf with other products"
)

for (i in seq_along(variations)) {
  foundry_image(
    paste("A blue ceramic coffee mug", variations[i]),
    model = "my-dalle",
    style = "natural"
  ) |>
    foundry_save_image(paste0("mug_", i, ".png"))
}
```

## Best Practices

1. **Save images promptly**: URLs expire quickly. Always save images you want to keep.

2. **Use HD quality sparingly**: HD takes longer and costs more. Use standard quality for drafts and iterations.

3. **Respect content policies**: DALL-E has content filters. Prompts requesting harmful, illegal, or inappropriate content will be rejected.

4. **Monitor costs**: Image generation is more expensive than text APIs. Track usage in the Azure Portal.

5. **Consider copyright**: AI-generated images have complex intellectual property implications. Consult legal guidance for commercial use.

## Next Steps

- Learn about [Content Safety](content-safety.html) for responsible AI deployment
- Explore [Text Embeddings](embeddings.html) for semantic search
- Read [Getting Started](getting-started.html) for foundryR basics
