---
title: "Image and Video Generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Image and Video Generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(foundryR)
```

foundryR supports current v1 preview image generation and editing parameters,
while keeping the legacy deployment-style image endpoint available with
`api = "deployment"`.

## Generate images

```{r image-call, eval = FALSE}
image <- foundry_image(
  "A clean data science illustration of survey responses becoming a chart",
  model = "gpt-image-1",
  size = "1024x1024",
  output_format = "png",
  background = "opaque"
)
```

```{r image-output}
image <- tibble::tibble(
  prompt = "A clean data science illustration of survey responses becoming a chart",
  revised_prompt = NA_character_,
  url = NA_character_,
  b64_json = "<base64 image bytes>",
  output_format = "png",
  created = as.POSIXct("2026-06-24 12:00:00", tz = "UTC")
)

image[, c("prompt", "url", "b64_json", "output_format", "created")]
```

Save generated image bytes with `foundry_save_image()`:

```{r save-image, eval = FALSE}
foundry_save_image(image, "survey-chart.png")
```

## Edit images

```{r edit-call, eval = FALSE}
edited <- foundry_image_edit(
  image = "survey-chart.png",
  prompt = "Use a blue Microsoft-inspired color palette.",
  model = "gpt-image-1",
  output_format = "png"
)
```

```{r edit-output}
edited <- tibble::tibble(
  prompt = "Use a blue Microsoft-inspired color palette.",
  revised_prompt = NA_character_,
  url = NA_character_,
  b64_json = "<base64 edited image bytes>",
  output_format = "png",
  created = as.POSIXct("2026-06-24 12:05:00", tz = "UTC")
)

edited[, c("prompt", "b64_json", "output_format", "created")]
```

## Start a video generation job

Video generation is a preview, long-running workflow. Create a job, poll it, and
download content after a generation succeeds.

```{r video-call, eval = FALSE}
job <- foundry_video_job_create(
  "A short animation of dots clustering into groups",
  model = "my-video-model",
  width = 1280,
  height = 720,
  n_seconds = 5
)

foundry_video_job_get(job$job_id)
```

```{r video-output}
job <- tibble::tibble(
  job_id = "videojob_abc123",
  status = "running",
  model = "my-video-model",
  prompt = "A short animation of dots clustering into groups",
  width = 1280L,
  height = 720L,
  n_seconds = 5L,
  n_variants = 1L,
  created_at = as.POSIXct("2026-06-24 12:10:00", tz = "UTC")
)

job
```

```{r video-download, eval = FALSE}
foundry_video_download(
  generation_id = "vidgen_abc123",
  path = "cluster-animation.mp4"
)
```

## When to use these APIs

- Generate research stimuli or report illustrations with `foundry_image()`.
- Iterate on existing figures or image stimuli with `foundry_image_edit()`.
- Treat video as experimental: keep prompts, job IDs, and output files together
  so generated assets remain reproducible and auditable.
