---
title: "Responses API, Structured Extraction, and Web Search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Responses API, Structured Extraction, and Web Search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
fixture_dir <- "responses-api"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
run_api <- recording || have_fixtures

if (run_api) {
  httptest2::start_vignette(fixture_dir)
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_api
)
```

## Why the Responses API matters

Microsoft Foundry now exposes a newer v1 data-plane endpoint for Azure OpenAI:

```text
https://<resource>.openai.azure.com/openai/v1/responses
```

Unlike the older deployment-path chat API, the v1 Responses API sends the model
deployment in the JSON body. It also adds stateful response chaining, built-in
tools, structured output formats, and richer output metadata.

foundryR wraps this with `foundry_response()` while keeping the package's tidy
interface: generated text, citations, tool calls, token usage, and the raw
response are returned as tibble columns.

## Basic response

```{r basic-response}
library(foundryR)

foundry_response(
  "Answer in one sentence: what is retrieval-augmented generation?",
  model = "gpt-4.1"
)
```

The result includes:

- `response_id`: the stored Responses API object ID
- `output_text`: generated text aggregated from the response output items
- `citations`: a list-column of URL citations, when present
- `tool_calls`: a list-column of tool calls, such as web-search calls
- token usage columns, including reasoning and cached input tokens when the API
  reports them
- a `raw_response` list-column

## Stateful turns

Responses are stored by the service by default. You can chain turns by passing
the previous `response_id`:

```{r stateful-turns}
first <- foundry_response(
  "Define catastrophic forgetting in one sentence.",
  model = "gpt-4.1"
)

second <- foundry_response(
  "Explain it for a college freshman in one sentence.",
  model = "gpt-4.1",
  previous_response_id = first$response_id
)

second$output_text
```

If you do not want the service to store a response, pass `store = FALSE`.
Stateful chaining with `previous_response_id` requires the previous response to
be stored.

## Structured extraction with JSON Schema

`foundry_extract()` is designed for data scientists and researchers who need to
turn free text into analyzable variables. You provide a JSON Schema as an R
list; foundryR sends it through the Responses API structured output format and
returns one row per input text.

`foundry_extract()` sends `strict = TRUE` in the JSON Schema format by default.
For supported models, the service must return data that conforms to the schema.

```{r structured-extraction}
schema <- list(
  type = "object",
  properties = list(
    sentiment = list(
      type = "string",
      enum = c("positive", "negative", "neutral")
    ),
    entities = list(
      type = "array",
      items = list(type = "string")
    ),
    summary = list(type = "string")
  ),
  required = c("sentiment", "entities", "summary"),
  additionalProperties = FALSE
)

texts <- c(
  "The new data pipeline reduced manual coding time by half.",
  "Participants reported confusion about the consent form."
)

foundry_extract(
  texts,
  schema = schema,
  model = "gpt-4.1"
)
```

Top-level scalar fields become regular tibble columns. Arrays and nested objects
become list-columns, which work naturally with tidyverse workflows.

## User-defined R tools

The Responses API function-calling contract uses tool definitions with
`type = "function"` and follow-up tool outputs with
`type = "function_call_output"` plus a matching `call_id`. `foundry_tool()`
builds the tool schema and keeps the R function reference for local execution.
`foundry_agent()` runs the bounded call, execute, return-output loop.

```{r function-tools, eval = FALSE}
get_weather <- function(location) {
  list(location = location, temperature = "70 F")
}

weather_tool <- foundry_tool(
  get_weather,
  description = "Get weather for a location",
  parameters = list(
    type = "object",
    properties = list(location = list(type = "string")),
    required = "location"
  )
)

turns <- foundry_agent(
  "What is the weather in San Francisco?",
  tools = list(weather_tool),
  model = "gpt-4.1",
  max_iterations = 4
)

turns[, c("iteration", "final", "output_text")]
turns$tool_results[[1]]
```

The loop stops with an error if the model continues requesting tools after
`max_iterations`. This protects batch jobs from unbounded tool use.

## Remote MCP tools

Microsoft documents remote Model Context Protocol tools for the Responses API.
foundryR does not add a separate MCP helper yet because `foundry_response()`
already accepts raw Responses API tool objects:

```{r mcp-tool, eval = FALSE}
mcp_tool <- list(
  type = "mcp",
  server_label = "my_mcp_server",
  server_url = Sys.getenv("MY_MCP_SERVER_URL"),
  require_approval = "never"
)

foundry_response(
  "Use the MCP server if it helps answer the question.",
  model = "gpt-4.1",
  tools = list(mcp_tool)
)
```

Only attach MCP servers you trust and whose data-handling behavior your
organization has approved.

## Web-grounded answers with citations

`foundry_web_search()` uses the Responses API `web_search` tool and parses URL
citations into a tidy list-column:

```{r web-search, eval = FALSE}
answer <- foundry_web_search(
  "What changed recently in Azure AI Foundry Responses API?",
  model = "gpt-4.1",
  search_context_size = "high"
)

answer$output_text
answer$citations[[1]]
answer$tool_calls[[1]]
```

You can optionally provide approximate location fields:

```{r web-search-location, eval = FALSE}
foundry_web_search(
  "Find a recent AI research event near me.",
  model = "gpt-4.1",
  country = "US",
  region = "Washington",
  city = "Seattle",
  timezone = "America/Los_Angeles"
)
```

## Responsible use of web search

Microsoft documents that web search uses Grounding with Bing Search and/or
Grounding with Bing Custom Search. The Data Protection Addendum does not apply
to data sent to these services, data can leave compliance and geographic
boundaries, and tool usage can incur additional costs. Avoid sending secrets or
sensitive research data in web-search prompts.

## Reasoning models and token accounting

`foundry_response()` accepts `reasoning_effort` and sends it as
`reasoning = list(effort = ...)`, the Responses API shape documented by
Microsoft for reasoning models. `foundry_chat()` accepts the chat-completions
shape, `reasoning_effort = "medium"`.

```{r reasoning, eval = FALSE}
foundry_response(
  "Compare the two arguments and identify the weaker premise.",
  model = "my-reasoning-deployment",
  reasoning_effort = "medium"
)
```

The returned tibble includes `reasoning_tokens` and `cached_input_tokens` when
the API reports them. These fields matter for cost review because reasoning
tokens may be billed even when they are not visible in `output_text`.

## Streaming

The Azure OpenAI Responses API supports Server-Sent Events streaming, but
foundryR does not implement streaming. The package focuses on reproducible,
tibble-returning analytical workflows. Use ellmer when you need interactive
streaming chat in R.

## When to use `foundry_chat()` vs `foundry_response()`

Use `foundry_chat()` when you want the established chat-completions interface
and simple assistant replies.

Use `foundry_response()` when you need newer v1 capabilities: stateful response
IDs, built-in tools, structured output formats, richer output items, or a
forward-looking API surface for new Microsoft Foundry model capabilities.

```{r cleanup, include = FALSE}
if (run_api) {
  httptest2::end_vignette()
}
```
