---
title: "tidymodels Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidymodels Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

foundryR integrates seamlessly with the [tidymodels](https://www.tidymodels.org/) ecosystem through `step_foundry_embed()`, a recipe step that converts text columns into embedding vectors. This enables you to incorporate state-of-the-art text representations into your machine learning pipelines alongside traditional preprocessing steps.

## Why Use Embeddings in ML Pipelines?

Traditional text features like bag-of-words or TF-IDF capture word frequencies but miss semantic meaning. Embeddings provide dense vector representations that understand:

- **Synonyms**: "happy" and "joyful" produce similar vectors
- **Context**: "bank" (financial) vs "bank" (river) are distinguished
- **Relationships**: Semantic similarities are preserved in vector space

By converting text to embeddings within a recipe, you get:

- **Reproducible preprocessing**: The embedding step is part of your documented workflow
- **Consistent handling**: Training and test data are processed identically
- **Pipeline integration**: Combine with other preprocessing steps seamlessly

## Prerequisites

Install tidymodels if you haven't already:

```{r install}
install.packages("tidymodels")
```

Ensure foundryR is configured with your Azure credentials and you have an embedding model deployed.

## Basic Usage

### Creating a Recipe with Embeddings

Use `step_foundry_embed()` to add embedding generation to your recipe:

```{r basic-recipe}
library(tidymodels)
library(foundryR)

# Sample data with text and outcome
reviews <- tibble(
  text = c(
    "This product is amazing, highly recommend!",
    "Terrible quality, waste of money",
    "Good value for the price",
    "Disappointed with the purchase",
    "Exceeded my expectations",
    "Would not buy again"
  ),
  sentiment = factor(c("positive", "negative", "positive",
                       "negative", "positive", "negative"))
)

# Create recipe with embedding step
recipe_spec <- recipe(sentiment ~ text, data = reviews) %>%
  step_foundry_embed(
    text,
    model = "text-embedding-3-small",  # Your deployment name
    keep_original = FALSE         # Remove original text column
  )

recipe_spec
#> Recipe
#>
#> Inputs:
#>
#>       role #variables
#>    outcome          1
#>  predictor          1
#>
#> Operations:
#>
#> Azure AI Foundry embeddings for text
```

### Preparing and Baking the Recipe

```{r prep-bake}
# Prepare the recipe (generates embeddings for training data)
prepped_recipe <- prep(recipe_spec, training = reviews)

# Bake to transform data
baked_data <- bake(prepped_recipe, new_data = NULL)
baked_data
#> # A tibble: 6 × 1,537
#>   sentiment text_embed_001 text_embed_002 text_embed_003 ... text_embed_1536
#>   <fct>              <dbl>          <dbl>          <dbl>             <dbl>
#> 1 positive          0.0234        -0.0156         0.0089           0.0123
#> 2 negative         -0.0145         0.0234        -0.0067          -0.0089
#> 3 positive          0.0178        -0.0123         0.0145           0.0098
#> ...
```

The text column is replaced with 1,536 numeric embedding dimensions (the exact number depends on your embedding model).

## Complete ML Pipeline Example

Here's a full example building a sentiment classifier:

```{r full-pipeline}
library(tidymodels)
library(foundryR)

# Load your data
set.seed(123)
reviews <- tibble(
  review_text = c(
    # Positive reviews
    "Absolutely love this product! Works perfectly.",
    "Great quality and fast shipping. Very satisfied.",
    "Best purchase I've made this year. Highly recommend!",
    "Exceeded all expectations. Will buy again.",
    "Perfect fit and great value for money.",
    # Negative reviews
    "Complete waste of money. Broke after one use.",
    "Terrible customer service. Never buying again.",
    "Poor quality, doesn't work as advertised.",
    "Disappointed. Much smaller than expected.",
    "Arrived damaged and took forever to ship."
  ),
  sentiment = factor(rep(c("positive", "negative"), each = 5))
)

# Split data
splits <- initial_split(reviews, prop = 0.8, strata = sentiment)
train_data <- training(splits)
test_data <- testing(splits)

# Define recipe with embeddings
embedding_recipe <- recipe(sentiment ~ review_text, data = train_data) %>%
  step_foundry_embed(
    review_text,
    model = "text-embedding-3-small",
    keep_original = FALSE
  ) %>%
  step_normalize(all_numeric_predictors())  # Normalize embedding dimensions

# Define model
log_reg_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Create workflow
sentiment_workflow <- workflow() %>%
  add_recipe(embedding_recipe) %>%
  add_model(log_reg_spec)

# Fit the model
fitted_workflow <- fit(sentiment_workflow, data = train_data)

# Make predictions on test data
predictions <- predict(fitted_workflow, test_data) %>%
  bind_cols(test_data)

# Evaluate
predictions %>%
  metrics(truth = sentiment, estimate = .pred_class)
#> # A tibble: 2 × 3
#>   .metric  .estimator .estimate
#>   <chr>    <chr>          <dbl>
#> 1 accuracy binary         1
#> 2 kap      binary         1
```

## Advanced Options

### Controlling Embedding Dimensions

Some models support dimension reduction for faster processing:

```{r dimensions}
recipe_spec <- recipe(sentiment ~ text, data = reviews) %>%
  step_foundry_embed(
    text,
    model = "text-embedding-3-small",
    dimensions = 256,  # Reduce from 1536 to 256
    keep_original = FALSE
  )
```

Lower dimensions mean:
- Faster model training
- Less memory usage
- Some loss in semantic precision

### Multiple Text Columns

Process multiple text columns independently:

```{r multi-column}
# Data with multiple text fields
data <- tibble(
  title = c("Great Product", "Terrible Experience"),
  description = c("Works as expected", "Broke immediately"),
  outcome = c(1, 0)
)

recipe_spec <- recipe(outcome ~ ., data = data) %>%
  step_foundry_embed(title, model = "text-embedding-3-small",
                     prefix = "title_") %>%
  step_foundry_embed(description, model = "text-embedding-3-small",
                     prefix = "desc_") %>%
  step_rm(title, description)  # Remove original text columns
```

### Keeping Original Columns

Sometimes you want both the text and embeddings:

```{r keep-original}
recipe_spec <- recipe(sentiment ~ text, data = reviews) %>%
  step_foundry_embed(
    text,
    model = "text-embedding-3-small",
    keep_original = TRUE  # Keep the text column
  )

# Useful when you also want to apply other text processing
```

### Custom Column Prefix

Control the naming of embedding columns:

```{r prefix}
recipe_spec <- recipe(sentiment ~ text, data = reviews) %>%
  step_foundry_embed(
    text,
    model = "text-embedding-3-small",
    prefix = "embed_"  # Columns will be embed_001, embed_002, etc.
  )
```

## Cross-Validation

Embeddings are generated during `prep()`, so cross-validation works correctly:

```{r cv}
# Create CV folds
folds <- vfold_cv(train_data, v = 5, strata = sentiment)

# Fit resamples
cv_results <- fit_resamples(
  sentiment_workflow,
  resamples = folds,
  metrics = metric_set(accuracy, roc_auc)
)

# Collect metrics
collect_metrics(cv_results)
#> # A tibble: 2 × 6
#>   .metric  .estimator  mean     n std_err .config
#>   <chr>    <chr>      <dbl> <int>   <dbl> <chr>
#> 1 accuracy binary     0.875     5  0.0559 Preprocessor1_Model1
#> 2 roc_auc  binary     0.925     5  0.0433 Preprocessor1_Model1
```

## Hyperparameter Tuning

Tune the embedding dimensions alongside model hyperparameters:
```{r tuning}
# Recipe with tunable dimensions
tunable_recipe <- recipe(sentiment ~ text, data = train_data) %>%
  step_foundry_embed(
    text,
    model = "text-embedding-3-small",
    dimensions = tune(),  # Will be tuned
    keep_original = FALSE
  ) %>%
  step_normalize(all_numeric_predictors())

# Model with tunable parameters
rf_spec <- rand_forest(
  mtry = tune(),
  trees = 500,
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Workflow
tunable_workflow <- workflow() %>%
  add_recipe(tunable_recipe) %>%
  add_model(rf_spec)

# Define grid
grid <- grid_regular(
  dimensions(range = c(128, 512)),  # Embedding dimensions
  mtry(range = c(10, 50)),
  min_n(range = c(2, 10)),
  levels = 3
)

# Tune (this will take a while due to API calls)
tune_results <- tune_grid(
  tunable_workflow,
  resamples = folds,
  grid = grid,
  metrics = metric_set(accuracy, roc_auc)
)

# Best parameters
show_best(tune_results, metric = "roc_auc")
```

## Performance Considerations

### API Rate Limits

Embedding generation makes API calls for each text. For large datasets:

1. **Use batch processing**: Consider using `foundry_embed_batch()` outside the recipe for large training sets
2. **Cache embeddings**: Pre-compute and store embeddings for frequently used datasets
3. **Reduce cross-validation folds**: Fewer folds mean fewer API calls during tuning

### Cost Management

Each embedding call incurs API costs. Strategies to manage costs:

- Start with smaller dimension sizes during development
- Use a subset of data for initial experimentation
- Pre-compute embeddings for production datasets

### Memory Usage

With 1,536 dimensions per text and thousands of observations, memory can grow quickly:

```{r memory}
# Estimate memory for 10,000 texts
n_texts <- 10000
n_dims <- 1536
bytes_per_double <- 8

memory_mb <- (n_texts * n_dims * bytes_per_double) / 1024^2
print(paste(round(memory_mb), "MB for embeddings alone"))
#> [1] "117 MB for embeddings alone"
```

Consider dimension reduction for large datasets.

## Troubleshooting

### "Column already exists" Error

If you run `prep()` multiple times, column names may conflict:

```{r troubleshoot}
# Use a unique prefix if reusing recipes
recipe_spec <- recipe(sentiment ~ text, data = reviews) %>%
  step_foundry_embed(text, model = "my-model",
                     prefix = paste0("v", format(Sys.time(), "%H%M%S"), "_"))
```

### Rate Limit Errors

If you hit rate limits during prep:

```{r rate-limit}
# Prepare in smaller batches
small_sample <- reviews %>% slice_sample(n = 100)
prepped <- prep(recipe_spec, training = small_sample)
```

### Missing Credentials

Ensure credentials are set before creating recipes:

```{r credentials}
# Check setup
foundry_check_setup()

# Set credentials if needed
foundry_set_endpoint("AZURE_FOUNDRY_ENDPOINT")
foundry_set_key("your-api-key")
```

## Next Steps

- Learn about [Text Embeddings](embeddings.html) in depth
- Explore [Content Safety](content-safety.html) for responsible AI
- Read the [tidymodels documentation](https://www.tidymodels.org/) for more preprocessing options
