---
title: "Integrating foundryR with onet2r"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
# This article runs against recorded, credential-free API fixtures. When the
# fixtures are present (recorded once with data-raw/record-doc-outputs.R and
# committed under vignettes/articles/onet2r-integration/), every onet_*() and
# foundry_*() call below is executed and its real output is shown. When they are
# absent -- or when onet2r is not installed -- the API chunks are not evaluated,
# so the article still builds anywhere with no O*NET or Azure credentials.
#
# onet2r is a GitHub-only package, which is why this is a pkgdown article rather
# than a shipped vignette: the article documents the integration on the website
# without adding a non-CRAN dependency to the package build.
fixture_dir <- "onet2r-integration"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
have_onet2r <- requireNamespace("onet2r", quietly = TRUE)
run_api <- recording || have_fixtures
can_run <- run_api && have_onet2r

if (run_api) {
  httptest2::start_vignette(fixture_dir)
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = can_run
)
```

## Why combine the two packages

[onet2r](https://github.com/farach/onet2r) reads the U.S. Department of Labor's
O\*NET database: occupation titles, descriptions, skills, tasks, and technology
requirements for roughly a thousand occupations. It returns tidy tibbles.

foundryR turns text into data with Azure AI Foundry: embeddings for semantic
comparison, and chat completions for summarization. The two fit together
naturally because onet2r produces the text that foundryR reasons over, and both
speak tibbles.

This article builds one end-to-end workflow: pull real occupations from O\*NET,
embed their titles, and rank them against a plain-language query by meaning
rather than by keyword. It closes by summarizing the top match's real O\*NET
description with a chat model.

## Setup

onet2r is on GitHub, not CRAN. Install both packages with pak:

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("farach/foundryR")
pak::pak("farach/onet2r")
```

Each package reads its own credentials from the environment, so nothing secret
appears in your code:

```{r credentials, eval = FALSE}
# Azure AI Foundry (foundryR)
foundry_set_endpoint(Sys.getenv("AZURE_FOUNDRY_ENDPOINT"))
foundry_set_key(Sys.getenv("AZURE_FOUNDRY_KEY"))

# O*NET (onet2r) reads ONET_API_KEY. Register for a free key at
# https://services.onetcenter.org/developer/ then set:
Sys.setenv(ONET_API_KEY = "your-onet-key")
```

```{r libraries}
library(foundryR)
library(onet2r)
library(dplyr)
```

## Reading occupation data from O\*NET

`onet_search()` matches occupations by keyword and returns a tibble of `code`
and `title`:

```{r onet-search}
security_matches <- onet_search("information security")
security_matches
```

`onet_occupation()` returns the full record for one occupation code as a list.
The `description` field is the plain-language summary O\*NET writes for each job:

```{r onet-occupation}
analyst <- onet_occupation("15-1212.00")
analyst$title
substr(analyst$description, 1, 220)
```

## Building a semantic search index

Keyword search only finds occupations whose titles contain the words you typed.
Embeddings find occupations by meaning. Start by pulling a block of occupations
and embedding their titles.

`onet_occupations()` lists occupations in O\*NET-SOC code order; the first 150
codes span management, business, and computer occupations.

```{r corpus}
occupations <- onet_occupations(end = 150)

title_embeddings <- foundry_embed(
  occupations$title,
  model = "text-embedding-3-small"
)

occupation_index <- occupations |>
  mutate(embedding = title_embeddings$embedding)

occupation_index
```

`foundry_embed()` returns one row per input with the vector in a list-column, so
attaching it back onto the occupation tibble keeps everything in one frame.

## Searching by meaning

Embed a natural-language query the same way, then rank every occupation by cosine
similarity to it. The query below shares no words with the official titles it
should surface.

```{r semantic-search}
query <- "protecting company networks and data from hackers and cyber attacks"

query_vec <- foundry_embed(query, model = "text-embedding-3-small")$embedding[[1]]

cosine <- function(a, b) sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))

ranked <- occupation_index |>
  mutate(similarity = vapply(embedding, cosine, numeric(1), b = query_vec)) |>
  arrange(desc(similarity)) |>
  select(code, title, similarity)

head(ranked, 5)
```

The top results are the security- and network-focused occupations, even though
the query never uses their title words. That is the advantage of embeddings over
keyword lookup: "protecting data from hackers" resolves to "Information Security
Analysts" on meaning alone.

## Inspecting skills for a match

Skill, knowledge, and task endpoints each return a tidy tibble, so they drop
straight into a dplyr pipeline. Here are the first rows of the skills table for
the top match:

```{r skills}
top_code <- ranked$code[1]

onet_skills(top_code) |>
  head()
```

## Summarizing the match with a chat model

The pieces combine cleanly: take the real O\*NET description for the top match and
ask a chat model to rewrite it for a specific audience. `foundry_chat()` takes the
prompt as `message` and returns a tibble; the generated text is in `content`.

```{r ai-summary}
top_occupation <- onet_occupation(top_code)

summary <- foundry_chat(
  message = paste(
    "Summarize this occupation for someone considering a career change,",
    "in two sentences:\n\n",
    top_occupation$description
  )
)

cat(summary$content)
```

## Where to take it

This workflow generalizes past a single query. Embed the whole occupation index
once, cache it, and you have a reusable semantic job-matcher: score a resume or a
free-text career interest against every occupation, cluster occupations by skill
profile, or flag near-duplicate roles. onet2r supplies the authoritative text and
foundryR turns it into comparable numbers and readable summaries -- all in
tibbles.

```{r cleanup, include = FALSE}
if (run_api) {
  httptest2::end_vignette()
}
```
