---
title: "Match job descriptions to O*NET occupations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Match job descriptions to O*NET occupations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# onet2r is on GitHub, not CRAN, so this is a pkgdown article rather than a
# shipped vignette. Without onet2r or recorded fixtures the article still builds,
# showing code only.
fixture_dir <- "onet2r-integration"
recording <- nzchar(Sys.getenv("FOUNDRY_RECORD_DOCS"))
have_fixtures <- dir.exists(fixture_dir) && length(list.files(fixture_dir)) > 0
have_onet2r <- requireNamespace("onet2r", quietly = TRUE)
run_api <- requireNamespace("httptest2", quietly = TRUE) && have_onet2r &&
  (recording || have_fixtures)
library(foundryR)
if (run_api) {
  httptest2::start_vignette(fixture_dir)
}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = run_api)
```

Surveys, job postings, and administrative records often describe work in free text: "I fix the office computers", "I teach fourth grade". Before those records can be joined to wages, skills, or task measures, each one needs a standard occupation code. This article matches short job descriptions to O\*NET-SOC occupations with foundryR embeddings, checks the matches against hand codes, and pulls the skills profile for a matched occupation with [onet2r](https://github.com/farach/onet2r).

Calls to O\*NET and Azure show output recorded from a live run; setup code is shown but not run.

## Setup

onet2r reads O\*NET Web Services and archived O\*NET database releases into tibbles. It is on GitHub, not CRAN.

```{r install, eval = FALSE}
install.packages("foundryR")

# install.packages("pak")
pak::pak("farach/onet2r")
```

Both packages read credentials from the environment. onet2r needs a free key from the [O\*NET Web Services developer site](https://services.onetcenter.org/developer/), stored in `.Renviron` as `ONET_API_KEY`.

```{r credentials, eval = FALSE}
# In .Renviron:
# ONET_API_KEY=your-onet-key

foundry_set_endpoint(Sys.getenv("AZURE_FOUNDRY_ENDPOINT"))
foundry_set_key(Sys.getenv("AZURE_FOUNDRY_KEY"))
```

```{r libraries, message = FALSE}
library(foundryR)
library(onet2r)
library(dplyr)
```

## Get the candidate occupations

`onet_occupations_all()` returns the code and title of every occupation in the current O\*NET-SOC taxonomy in one call.

```{r catalog}
catalog <- onet_occupations_all(show_progress = FALSE)
nrow(catalog)
```

This example compares the descriptions with four major groups: computer and mathematical (15), education (25), healthcare practitioners (29), and office support (43). The smaller candidate set keeps the recorded example small. In a study, embed the full catalog once and store the vectors.

```{r candidates}
candidates <- catalog |>
  filter(substr(code, 1, 2) %in% c("15", "25", "29", "43"))

count(candidates, major_group = substr(code, 1, 2))
```

## Write down the hand codes first

The descriptions below were written for this article, each with the code a person would assign. In a study, draw the validation sample at random from your data and have people code it before they see any model matches.

```{r responses, eval = TRUE}
responses <- tibble::tribble(
  ~id,  ~job_text,                                                                  ~hand_code,
  "r1", "I keep the office computers running, fix printers, and reset passwords.",  "15-1232.00",
  "r2", "I watch our network for intrusions and patch servers after alerts.",       "15-1212.00",
  "r3", "I teach fourth graders reading and math.",                                 "25-2021.00",
  "r4", "I teach statistics courses at a community college.",                       "25-1022.00",
  "r5", "I care for patients on a hospital ward, give medications, and chart vitals.", "29-1141.00",
  "r6", "I clean teeth and take X-rays at a dental clinic.",                        "29-1292.00",
  "r7", "I answer phones, greet visitors, and book appointments at the front desk.", "43-4171.00",
  "r8", "I process invoices and reconcile vendor payments for the accounting team.", "43-3031.00"
)
```

Check that every hand code is a current O\*NET-SOC code. A code from an older taxonomy can never match, and the miss would look like a model error.

```{r code-check}
setdiff(responses$hand_code, candidates$code)
```

## Embed both sides with the same settings

Embeddings place texts with similar meanings near each other. Embed the occupation titles and the job descriptions with the same model and the same `dimensions`; vectors from different settings are not comparable. Shorter vectors keep the stored matrix small.

```{r embed}
title_vectors <- foundry_embed(
  candidates$title,
  model = "text-embedding-3-small",
  dimensions = 256
)
job_vectors <- foundry_embed(
  responses$job_text,
  model = "text-embedding-3-small",
  dimensions = 256
)

c(titles_failed = sum(title_vectors$.error), jobs_failed = sum(job_vectors$.error))
```

## Rank occupations for each description

Dividing each vector by its length turns a matrix product into a table of cosine similarities, with one row per description and one column per occupation. The five highest scores in each row form a shortlist.

```{r rank}
unit_rows <- function(x) {
  m <- do.call(rbind, x$embedding)
  m / sqrt(rowSums(m^2))
}
similarity <- unit_rows(job_vectors) %*% t(unit_rows(title_vectors))

shortlist <- purrr::map_dfr(seq_len(nrow(responses)), function(i) {
  best <- order(similarity[i, ], decreasing = TRUE)[1:5]
  tibble::tibble(
    id = responses$id[[i]],
    rank = 1:5,
    code = candidates$code[best],
    title = candidates$title[best],
    similarity = similarity[i, best]
  )
})

first_choice <- shortlist |>
  filter(rank == 1) |>
  left_join(select(responses, id, hand_code), by = "id") |>
  mutate(agrees = code == hand_code)

first_choice |>
  select(id, title, similarity, agrees)
```

## Check the first choices against the hand codes

An O\*NET-SOC code has eight digits. The first six are the federal SOC code, which wage and employment data such as BLS OEWS use, and the last two pick out a more detailed O\*NET occupation. A match can miss the detailed occupation but land in the right SOC occupation, so count agreement at both levels, along with how often the hand code makes the shortlist.

```{r agreement}
soc <- function(code) substr(code, 1, 7)

on_shortlist <- shortlist |>
  left_join(select(responses, id, hand_code), by = "id") |>
  group_by(id) |>
  summarise(found = any(code == hand_code))

tibble::tibble(
  descriptions = nrow(responses),
  exact = sum(first_choice$agrees),
  same_soc = sum(soc(first_choice$code) == soc(first_choice$hand_code)),
  on_shortlist = sum(on_shortlist$found)
)
```

```{r agreement-summary, echo = FALSE, results = "asis"}
misses <- first_choice |>
  filter(!agrees) |>
  left_join(select(candidates, hand_code = code, hand_title = title), by = "hand_code")
cat(sprintf(
  "In this recording the first choice matches the hand code for %d of %d descriptions, and the hand code is on the shortlist for %d.",
  sum(first_choice$agrees), nrow(first_choice), sum(on_shortlist$found)
))
if (nrow(misses) > 0) {
  cat(sprintf(
    " The misses: %s.",
    paste(sprintf("%s matched %s instead of %s", misses$id, misses$title, misses$hand_title), collapse = "; ")
  ))
}
cat(" Titles are short, and many differ by a word or two, so title similarity alone is a weak final answer. It works better as a way to build a shortlist.\n")
```

## Let a model choose from the shortlist

A model can read the whole description and compare it with each candidate on the shortlist. The schema's enum lists every shortlisted code, and the instructions limit each answer to the candidates listed with that description.

```{r choose}
choice_input <- shortlist |>
  group_by(id) |>
  summarise(options = paste0("- ", code, ": ", title, collapse = "\n")) |>
  left_join(responses, by = "id") |>
  mutate(prompt = paste0("Job description: ", job_text, "\n\nCandidate occupations:\n", options))

choice_schema <- foundry_schema(
  choice = schema_enum(
    sort(unique(shortlist$code)),
    "Code of the candidate occupation that best matches the job description."
  )
)

chosen <- foundry_extract(
  choice_input,
  text_col = "prompt",
  schema = choice_schema,
  instructions = paste(
    "Pick the one candidate occupation that best matches the job description.",
    "Choose only from the candidates listed with the description."
  )
)

chosen |>
  left_join(select(candidates, choice = code, chosen_title = title), by = "choice") |>
  mutate(agrees = choice == hand_code) |>
  select(id, chosen_title, agrees, .error)
```

```{r choose-summary, echo = FALSE, results = "asis"}
listed <- purrr::map2_lgl(chosen$id, chosen$choice, function(i, code) {
  code %in% shortlist$code[shortlist$id == i]
})
model_misses <- chosen |>
  filter(choice != hand_code) |>
  left_join(select(candidates, choice = code, chosen_title = title), by = "choice") |>
  left_join(select(candidates, hand_code = code, hand_title = title), by = "hand_code") |>
  left_join(on_shortlist, by = "id")
cat(sprintf(
  "The model's choice matches the hand code for %d of %d descriptions, against %d of %d for the first embedding match.",
  sum(chosen$choice == chosen$hand_code, na.rm = TRUE), nrow(chosen),
  sum(first_choice$agrees), nrow(first_choice)
))
if (nrow(model_misses) > 0) {
  cat(sprintf(
    " The misses: %s.",
    paste(sprintf(
      "%s chose %s instead of %s%s",
      model_misses$id, model_misses$chosen_title, model_misses$hand_title,
      ifelse(model_misses$found, "", ", which was not on its shortlist")
    ), collapse = "; ")
  ))
  if (any(!model_misses$found)) {
    cat(" A model cannot pick a code the shortlist leaves out, so a longer shortlist or embeddings of the O\\*NET descriptions would give it a better set to choose from.")
  }
}
if (!all(listed, na.rm = TRUE)) {
  cat(sprintf(" %d choice(s) came from another description's shortlist, which the shared enum allows; filter those out or send them to a coder.", sum(!listed, na.rm = TRUE)))
}
cat("\n")
```

## Send disagreements to a person

The two methods fail in different ways, so a row where they disagree deserves a second look. Review those rows first, then spot-check a random sample of the rows where they agree.

```{r triage}
review <- chosen |>
  select(id, job_text, choice, hand_code) |>
  left_join(select(first_choice, id, embedding_first = code), by = "id") |>
  mutate(methods_disagree = choice != embedding_first)

review |>
  count(methods_disagree, correct = choice == hand_code)
```

```{r triage-summary, echo = FALSE, results = "asis"}
flagged <- sum(review$methods_disagree, na.rm = TRUE)
wrong <- review$choice != review$hand_code
cat(sprintf(
  "Here the methods disagree on %d of %d descriptions. Of the %d rows where the model's choice is wrong, %d %s among the flagged rows.\n",
  flagged, nrow(review), sum(wrong, na.rm = TRUE),
  sum(wrong & review$methods_disagree, na.rm = TRUE),
  if (sum(wrong & review$methods_disagree, na.rm = TRUE) == 1) "is" else "are"
))
```
## Pull the skills profile for a match

Once a description has a code, onet2r can bring in O\*NET's measures for that occupation. `onet_skills()` returns a code's skills with an importance score; here are the top five for the nursing description's hand code.

```{r skills}
nurse_code <- responses$hand_code[responses$id == "r5"]

onet_skills(nurse_code, end = 5) |>
  select(name, importance)
```

## Record what you used

O\*NET updates the database several times a year, and occupation codes can change when the taxonomy is revised. For a study, record the date you pulled the catalog, the O\*NET release it came from, the embedding model and `dimensions`, the candidate groups, and the triage rule. onet2r's archive functions, such as `onet_archive_download()` and `onet_panel()`, let you pin an O\*NET release for longitudinal work.

This article includes information from [O\*NET Web Services](https://services.onetcenter.org/) by the U.S. Department of Labor, Employment and Training Administration (USDOL/ETA), used under the [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) license. O\*NET&reg; is a trademark of USDOL/ETA.

```{r cleanup, include = FALSE, eval = TRUE}
if (run_api) {
  httptest2::end_vignette()
}
```